### 2014-10-26 report/cover.tex
report template

\newcommand{\logo}[0]{\vspace*{1.5cm}\begin{center}\includegraphics[scale=0.4,keepaspectratio=true]{images/logo/EPFL.pdf} \end{center}\hrule\vspace{1cm}}
\linespread{1.125}\selectfont
\newcommand{\project}[1]{\begin{center}\large{#1}\end{center}}
\renewcommand{\title}[1]{\vspace{0.01cm}\begin{center}\LARGE{#1}\end{center}\vspace{0.2cm}}
\renewcommand{\author}[1]{\begin{center}\Large{#1}\end{center}}
\newcommand{\department}[1]{\begin{center}\large{#1}\end{center}}
\renewcommand{\date}[2]{\begin{center}\normalsize{#1 #2}\end{center}}

\newcommand{\supervisor}[4]{\begin{center}\begin{normalsize}{\bf #1}\\#2\\#3\\#4\end{normalsize}\end{center}}

\thispagestyle{empty}

\logo

\project{Pattern Classification and Machine Learning}
\title{Project 1}
\author{Kevin Serrano, SC-MA3\\Youssef El Baba, SC-MA3\\ Alexandre Helfre, SC-MA3}
\department{École Polytechnique Fédérale de Lausanne}
\date{November 3, 2014}

\begin{center}
\begin{tabular}{cc}
\begin{tabular}{p{6cm}}
\supervisor{Professor}{Mohammad Emtiyaz Khan}{LCA4}{EPFL}

\end{tabular}
\end{tabular}
\end{center}

\clearpage \thispagestyle{empty} \cleardoublepage

### 2014-10-26 sigmoid.m
"sigmoid function used in logistic regression"

@@ -1,0 +1,6 @@
+function [ Y ] = sigmoid( X )
+
+Y = 1./(1+exp(-X));
+
+end
+

### 2014-10-26 computeCostLS.m
LSGD doesn't work

@@ -3,7 +3,7 @@ function L = computeCostLS( y, tX, beta )
 
 N = length(y);
 e = y - tX*beta;
-L = e'*e/(2*N);
+L = (e'*e)/(2*N);
 
 end
 

### 2014-10-26 computeGradientLS.m

@@ -3,7 +3,7 @@ function g = computeGradientLS( y, tX, beta )
 %   
 N = length(y);
 e = y - tX*beta;
-g = -tX'*e/N;
+g = -(1/N)*(tX'*e);
 
 end
 

### 2014-10-26 gradientDescent.m

@@ -22,6 +22,7 @@ for k = 1:maxIters
     oldBetaNorm = sqrt(sum((oldBeta).^2));
     
     % gradient descen update
+
     beta = beta - alpha*g;
     newBetaNorm = sqrt(sum((beta).^2));
     
@@ -31,6 +32,7 @@ for k = 1:maxIters
     if (abs(newBetaNorm-oldBetaNorm)/oldBetaNorm < 10^-3)
         break
     end
+
 
     % print current corst and beta
     disp(['cost: ' num2str(L) ' and beta values ' num2str(beta(:)')]);

### 2014-10-26 testCode.m

@@ -17,23 +17,23 @@ eps = 1e-2;
 
 % test least squares via gradient descent
 disp('least squares via gradient descent...');
-alpha = 1;
+alpha = 0.05;
 tBeta = leastSquaresGD(y, tX, alpha);
 assert(all(abs(tBeta - beta) < eps));
 disp('OK!');
 
 % testing least squares 
-disp('least squares...');
-tBeta = leastSquares(y, tX);
-assert(all(abs(tBeta - beta) < eps));
-disp('OK!');
+% disp('least squares...');
+% tBeta = leastSquares(y, tX);
+% assert(all(abs(tBeta - beta) < eps));
+% disp('OK!');
 
 % testing ridge regression
-disp('ridge regression...');
-lambda = 1e-5;
-tBeta = ridgeRegression(y, tX, lambda);
-assert(all(abs(tBeta - beta) < eps));
-disp('OK!');
+% disp('ridge regression...');
+% lambda = 1e-5;
+% tBeta = ridgeRegression(y, tX, lambda);
+% assert(all(abs(tBeta - beta) < eps));
+% disp('OK!');
 
 % generate binary data
 y = (y>0);
@@ -47,13 +47,13 @@ assert(sum(tY ~= y) / size(y,1) < 0.2);
 disp('OK!');
 
 % testing penalize logistic regression
-disp('penalized logistic regression...');
-alpha = 1e-3;
-lambda = 1e-2;
-tBeta = penLogisticRegression(y,tX,alpha,lambda);
-tY = 1.0 ./ (1.0 + exp(-tX * tBeta)) > 0.5;
-assert(sum(tY ~= y) / size(y,1) < 0.2);
-disp('OK!');
+% disp('penalized logistic regression...');
+% alpha = 1e-3;
+% lambda = 1e-2;
+% tBeta = penLogisticRegression(y,tX,alpha,lambda);
+% tY = 1.0 ./ (1.0 + exp(-tX * tBeta)) > 0.5;
+% assert(sum(tY ~= y) / size(y,1) < 0.2);
+% disp('OK!');
 
 % code for writing csv files
 tY = 1.0 ./ (1.0 + exp(-tX * tBeta));

### 2014-10-26 gradientDescent.m
fixed gradient descent iteration number issue, and modified alpha in testCode

@@ -2,6 +2,8 @@ function [ beta ] = gradientDescent(y, tX, maxIters, alpha, betaStart, lossFun, 
 
 beta = betaStart;
 
+minIters = maxIters/5;
+
 for k = 1:maxIters
     
 
@@ -19,20 +21,16 @@ for k = 1:maxIters
     g = gradFun( y, tX, beta ) + 2*lambda*tempBeta;
     
     oldBeta = beta;
-    oldBetaNorm = sqrt(sum((oldBeta).^2));
     
     % gradient descen update
-
     beta = beta - alpha*g;
-    newBetaNorm = sqrt(sum((beta).^2));
     
     % tests whether the descent has converged ()
     
     % when the norm of the beta changed by less than 1/1000, time to stop
-    if (abs(newBetaNorm-oldBetaNorm)/oldBetaNorm < 10^-3)
+    if (min(abs(beta-oldBeta)./oldBeta) < 10^-3) && (k > minIters)
         break
     end
-
 
     % print current corst and beta
     disp(['cost: ' num2str(L) ' and beta values ' num2str(beta(:)')]);

### 2014-10-26 testCode.m

@@ -23,17 +23,17 @@ assert(all(abs(tBeta - beta) < eps));
 disp('OK!');
 
 % testing least squares 
-% disp('least squares...');
-% tBeta = leastSquares(y, tX);
-% assert(all(abs(tBeta - beta) < eps));
-% disp('OK!');
+disp('least squares...');
+tBeta = leastSquares(y, tX);
+assert(all(abs(tBeta - beta) < eps));
+disp('OK!');
 
 % testing ridge regression
-% disp('ridge regression...');
-% lambda = 1e-5;
-% tBeta = ridgeRegression(y, tX, lambda);
-% assert(all(abs(tBeta - beta) < eps));
-% disp('OK!');
+disp('ridge regression...');
+lambda = 1e-5;
+tBeta = ridgeRegression(y, tX, lambda);
+assert(all(abs(tBeta - beta) < eps));
+disp('OK!');
 
 % generate binary data
 y = (y>0);
@@ -47,13 +47,13 @@ assert(sum(tY ~= y) / size(y,1) < 0.2);
 disp('OK!');
 
 % testing penalize logistic regression
-% disp('penalized logistic regression...');
-% alpha = 1e-3;
-% lambda = 1e-2;
-% tBeta = penLogisticRegression(y,tX,alpha,lambda);
-% tY = 1.0 ./ (1.0 + exp(-tX * tBeta)) > 0.5;
-% assert(sum(tY ~= y) / size(y,1) < 0.2);
-% disp('OK!');
+disp('penalized logistic regression...');
+alpha = 1e-3;
+lambda = 1e-2;
+tBeta = penLogisticRegression(y,tX,alpha,lambda);
+tY = 1.0 ./ (1.0 + exp(-tX * tBeta)) > 0.5;
+assert(sum(tY ~= y) / size(y,1) < 0.2);
+disp('OK!');
 
 % code for writing csv files
 tY = 1.0 ./ (1.0 + exp(-tX * tBeta));

### 2014-10-26 gradientDescent.m
add 1/N to penalization term

@@ -13,12 +13,12 @@ for k = 1:maxIters
     
     % computing the loss using the given function, plus norm squared of
     % beta * lambda as a regularization term
-    L = lossFun( y, tX, beta ) + lambda*tempBetaNorm^2;
+    L = lossFun( y, tX, beta ) + 1/2*1/length(y)*lambda*tempBetaNorm^2;
     
     
     % comuting the gradient using the given function, if regularizing then
     % the regularization term's derivative is 2*lambda*beta
-    g = gradFun( y, tX, beta ) + 2*lambda*tempBeta;
+    g = gradFun( y, tX, beta ) + 1/length(y)*lambda*tempBeta;
     
     oldBeta = beta;
     

### 2014-10-26 testCode.m
testCode

@@ -17,7 +17,7 @@ eps = 1e-2;
 
 % test least squares via gradient descent
 disp('least squares via gradient descent...');
-alpha = 0.05;
+alpha = 1;
 tBeta = leastSquaresGD(y, tX, alpha);
 assert(all(abs(tBeta - beta) < eps));
 disp('OK!');

### 2014-10-26 testCode.m
all Test OK

@@ -17,7 +17,7 @@ eps = 1e-2;
 
 % test least squares via gradient descent
 disp('least squares via gradient descent...');
-alpha = 1;
+alpha = 0.05;
 tBeta = leastSquaresGD(y, tX, alpha);
 assert(all(abs(tBeta - beta) < eps));
 disp('OK!');

### 2014-10-28 report/paper.tex
report template update

\documentclass{article} % For LaTeX2e
% We will use NIPS submission format
\usepackage{nips13submit_e,times}
% for hyperlinks
\usepackage{hyperref}
\usepackage{url}
% For figures
\usepackage{graphicx} 
\usepackage{subfigure} 
% math packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsopn}
\usepackage{ifthen}
\usepackage{natbib}

\title{Project-I by Group MexicoCity}

\author{
Kevin Serrano\\EPFL\\
\texttt{kevin.serrano@epfl.ch} \And Youssef El Baba\\EPFL\\
\texttt{youssef.baba@epfl.ch} \And Alexandre Helfre\\EPFL\\
\texttt{alexandre.helfre@epfl.ch}
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\nipsfinalcopy

\begin{document}

\maketitle

\begin{abstract}
In this report, we discuss our implementation and findings for the project-I. \textcolor{red}{Todo at the end}

\end{abstract}
\section{Five functions}
\begin{itemize}
\item \textit{leastSquaresGD(y,tX,alpha)} : Least squares using gradient descent, alpha is the step-size
\item \textit{leastSquares(y,tX)} : Least squares using normal equations
\item \textit{ridgeRegression(y,tX,lambda)} : Ridge regression using normal equations, lambda is the regularization coefficient
\item \textit{logisticRegression(y,tX,alpha)} : Logistic regression using gradient descent or Newton's method
\item \textit{penLogisticRegression(y,tX,alpha,lambda)} : Penalized logistic regression

\end{itemize}

\section{Summary}
In this report, we analyzed a regression dataset and found that ridge regression is a reasonable fit. We estimate that the average test error is 1.213 ($\pm$ 0.02). We tried some feature transformation and found that there is a small improvement giving us a test error of around 1.198 ($\pm$ 0.015). This improvement, however, is not significant.

\subsubsection*{Acknowledgments}

\subsubsection*{References}

\end{document}

### 2014-10-28 gradientDescent.m
added verbose option (currently disabled) for enabling cost and beta function output

@@ -1,4 +1,4 @@
-function [ beta ] = gradientDescent(y, tX, maxIters, alpha, betaStart, lossFun, gradFun, lambda)
+function [ beta ] = gradientDescent(y, tX, maxIters, alpha, betaStart, lossFun, gradFun, lambda, verbose)
 
 beta = betaStart;
 
@@ -32,12 +32,16 @@ for k = 1:maxIters
         break
     end
 
-    % print current corst and beta
-    disp(['cost: ' num2str(L) ' and beta values ' num2str(beta(:)')]);
+    if verbose
+        % print current corst and beta
+        disp(['cost: ' num2str(L) ' and beta values ' num2str(beta(:)')]);
+    end
 
 end
 
-    disp(' ')
-    disp(['finished descent after ' num2str(k) ' iterations'])
+    if verbose
+        disp(' ')
+        disp(['finished descent after ' num2str(k) ' iterations'])
+    end
 end
 

### 2014-10-28 leastSquaresGD.m

@@ -6,7 +6,7 @@ lambda = 0;
 betaStart = zeros(size(tX,2),1);
 
 beta = gradientDescent(y,tX, maxIters, alpha, betaStart, @computeCostLS, ...
-    @computeGradientLS,lambda);
+    @computeGradientLS,lambda, 0);
 
 end
 

### 2014-10-28 logisticRegression.m

@@ -8,7 +8,7 @@ maxIters = 2000;
 betaStart = zeros(size(tX, 2), 1);
 
 beta = gradientDescent(y, tX, maxIters, alpha, betaStart,...
-    @computeCostLogReg, @computeGradientLogReg, 0);
+    @computeCostLogReg, @computeGradientLogReg, 0, 0);
 
 end
 

### 2014-10-28 penLogisticRegression.m

@@ -8,6 +8,6 @@ maxIters = 2000;
 betaStart = zeros(size(tX, 2), 1);
 
 beta = gradientDescent(y, tX, maxIters, alpha, betaStart,...
-    @computeCostLogReg, @computeGradientLogReg, lambda);
+    @computeCostLogReg, @computeGradientLogReg, lambda, 0);
 
 end

### 2014-10-28 youssef plays/checkNormality.m
"added my own test/script files, please do the same"

@@ -1,0 +1,24 @@
+function [ normalityTests, pValues ] = checkNormality( results, noOfVariables )
+% testing the normal distribution hypothesis for each IQM (column), and also
+% for the success criteria (also used as variable in computing coefficients
+% in other functions) of the results, seperately for each pool (scenario,
+% real-world photo). Test based on the One-sample Kolmogorov-Smirnov method
+
+% initialization
+normalityTests = zeros(noOfVariables, 1);
+pValues = zeros(noOfVariables, 1);
+
+% normality tests computations for the IQMs
+for k = 1:noOfVariables
+    temp = results(:, k);
+    
+    if sum(isnan(temp)) == size(temp, 1)
+        normalityTests(k) = 0;
+        pValues(k) = NaN;
+    else
+        
+    [normalityTests(k), pValues(k)] = kstest(temp);
+    end
+end
+
+end

### 2014-10-28 youssef plays/computeCorrelationCoeffs.m

@@ -1,0 +1,36 @@
+function [ spearmanCoeffs, spearmanSig, pearsonCoeffs, pearsonSig,...
+    normalityTests, corrExistence, corrSuitability] = ...
+    computeCorrelationCoeffs( statistics, noOfVariables )
+% computes the correlation coefficient (both pearson & spearman) for each
+% input variable with respect to output variable
+
+% initialization of the normality test results
+normalityTests = zeros(noOfVariables, 2);
+
+
+% normality testing for each variable (columns) after pooling.
+[normalityTests(:, 1), normalityTests(:, 2)] =...
+    checkNormality( statistics, noOfVariables );
+
+
+% computation of IQM's spearman and pearson coefficients, using each
+% success criterion seperately
+[spearmanCoeffs, spearmanSig] = corr( statistics(:, 1), statistics(:, 2:end)...
+    ,'type', 'spearman', 'rows', 'complete');
+[pearsonCoeffs, pearsonSig] = corr( statistics(:, 1),...
+    statistics(:, 2:end), 'rows', 'complete');
+
+% deriving the hypothesis test results, 1 = relationship exists, 0 not
+spearmanSig = spearmanSig < 0.05;
+pearsonSig = pearsonSig < 0.05;
+
+% computation of whether the IQMs, under each scenario, have ANY
+% statistical relationship with the succes criteria
+corrExistence = spearmanSig | pearsonSig;
+
+% computation of which correlation method best suits each IQM under a
+% specific success criterion (useful only for simple scenario).
+corrSuitability = pearsonSig - spearmanSig;
+% - 1 means spearman preferred, 0 neither, +1 pearson preferred
+
+end

### 2014-10-28 youssef plays/youssefplaying.m

@@ -1,0 +1,124 @@
+% Comments and observations made by youssef
+
+clc
+clear all
+close all
+
+%load('MexicoCity_regression.mat')
+load('MexicoCity_classification.mat')
+
+N = length(y_train); % data size
+D = size(X_train, 2); % dimensionality
+
+%% plotting the histograms of the variables
+noOfVertPlots = floor(sqrt(D));
+noOfHorFigures = ceil(D/noOfVertPlots);
+
+figure;
+suptitle('histogram of all the input variables')
+for i = 1:D
+    subplot(noOfVertPlots, noOfHorFigures, i)
+    hist(X_train(:, i))
+    title(num2str(i))
+end
+
+figure;
+hist(y_train);
+title('hisogram of output variable')
+
+%% normalizing the features
+
+% for regression variables
+% normalize features (store the mean and variance)
+meanX = mean(X_train);
+
+% 42 first features seem to be gaussian distributed, can be normalized
+normInputVars = 1:33;
+
+% normalizing the input variables
+X_train(:, normInputVars) = X_train(:, normInputVars) - repmat(meanX(normInputVars), N, 1);
+stdX = std(X_train);
+X_train(:, normInputVars) = X_train(:, normInputVars)./repmat(stdX(:, normInputVars), N, 1);
+
+% normalizing the output variable
+y_train = (y_train-mean(y_train))/std(y_train);
+
+% Form (y,tX) to get regression data in matrix form
+tX = [ones(N,1) X_train(:, :)];
+
+%% correlation & normal distribution analysis
+
+statistics = [y_train X_train];
+
+[ spearmanCoeffs, spearmanSig, pearsonCoeffs, pearsonSig,...
+    normalityTests, corrExistence, corrSuitability] = ...
+    computeCorrelationCoeffs( statistics, D+1);
+
+normalityTests = circshift(normalityTests, -1);
+
+figure;
+stem(spearmanSig.*spearmanCoeffs)
+title('correlated variables with coeff, 0 otherwise, spearman')
+
+figure;
+stem(pearsonSig.*pearsonCoeffs)
+title('correlated variables with coeff, 0 otherwise, pearson')
+
+%% preliminary fitting (regression)
+
[…]

### 2014-10-28 gradientDescent.m
modified the gradient descent to detect better change in beta to know when to stop. Also enabled verbose mode

@@ -28,7 +28,7 @@ for k = 1:maxIters
     % tests whether the descent has converged ()
     
     % when the norm of the beta changed by less than 1/1000, time to stop
-    if (min(abs(beta-oldBeta)./oldBeta) < 10^-3) && (k > minIters)
+    if (max(abs(beta-oldBeta)./oldBeta) < eps) && (k > minIters)
         break
     end
 

### 2014-10-28 leastSquaresGD.m

@@ -6,7 +6,7 @@ lambda = 0;
 betaStart = zeros(size(tX,2),1);
 
 beta = gradientDescent(y,tX, maxIters, alpha, betaStart, @computeCostLS, ...
-    @computeGradientLS,lambda, 0);
+    @computeGradientLS,lambda, 1);
 
 end
 

### 2014-10-28 logisticRegression.m

@@ -8,7 +8,7 @@ maxIters = 2000;
 betaStart = zeros(size(tX, 2), 1);
 
 beta = gradientDescent(y, tX, maxIters, alpha, betaStart,...
-    @computeCostLogReg, @computeGradientLogReg, 0, 0);
+    @computeCostLogReg, @computeGradientLogReg, 0, 1);
 
 end
 

### 2014-10-28 penLogisticRegression.m

@@ -8,6 +8,6 @@ maxIters = 2000;
 betaStart = zeros(size(tX, 2), 1);
 
 beta = gradientDescent(y, tX, maxIters, alpha, betaStart,...
-    @computeCostLogReg, @computeGradientLogReg, lambda, 0);
+    @computeCostLogReg, @computeGradientLogReg, lambda, 1);
 
 end

### 2014-10-28 testCode.m

@@ -17,7 +17,7 @@ eps = 1e-2;
 
 % test least squares via gradient descent
 disp('least squares via gradient descent...');
-alpha = 1;
+alpha = 0.01;
 tBeta = leastSquaresGD(y, tX, alpha);
 assert(all(abs(tBeta - beta) < eps));
 disp('OK!');

### 2014-10-30 logodds.m
"logloss"

@@ -1,0 +1,22 @@
+function [lo] = logodds(y, p)
+%LOGODDS compute the log odds for an output vector y and a vector
+%probability p
+
+% Error if y and p have not the same length
+if (length(y) ~= length(p))
+    error(' y and p have not the length')
+end
+    
+% y must be a vector
+y = y(:);
+
+% p must be a vector
+p = p(:);
+
+% check for the length of y
+N = length(y);
+
+% compute the logLoss
+lo = -(1/N).*sum((y.*log(p)+(1-p).*log(1-p)));
+
+

### 2014-10-30 decideRemove.m
01 loss + decideRemove

@@ -1,0 +1,20 @@
+function rA = decideRemove( A )
+% Decide which variables to remove from the Dx(D+1) matrix A.
+% We can remove one input variable v if the correlation between the v and
+% the ouput is zero and that all the row of v in A is 0 (except v corr v)
+outputCulumn = size(A,2);
+maxCorr = max(A(:,outputCulumn));
+threshold = 0.10*maxCorr;
+for i = 1:size(A,1)
+   if(A(i,outputCulumn)< threshold) % Ouput corr is less thant the threshold
+       % check if each element of this row is < threshold
+       s = sum(A(i,:)< threshold);
+       if(s==(size(A,1)-1)) % if all element of the row (except in diagonal) are less than the threshold then remove variable
+           A(i,:) = [];
+           A(:,i) = [];
+       end
+   end
+end
+rA = A;
+end
+

### 2014-10-30 loss01.m

@@ -1,0 +1,8 @@
+function loss = loss01( y,haty )
+%LOSS01 Summary of this function goes here
+%   Detailed explanation goes here
+N = length(y);
+loss = 1/N * sum(y~=haty);
+
+end
+

### 2014-10-30 spearmanCorr.m
"Spearman Correlation Matrix"

@@ -1,0 +1,11 @@
+function [C] = spearmanCorr(tX, y)
+% Compute the Spearman correlation in matrix C containing the pairwise 
+% linear correlation coefficient between each pair of columns. The last
+% column of the matrix C is the correlation between the input and the
+% output
+
+C = corr(tX,'type','Spearman');
+C = [C corr(tX,y, 'type', 'Spearman')];
+
+end
+

### 2014-10-31 genericKCV.m
added generic cross-validation + results plotting/analysis. Change name of loggOdds to LogLoss

@@ -1,0 +1,74 @@
+function [errorTr, errorTe] = genericKCV( y, tX,...
+    fittingFunction, errorFunction, lambdas, K, alpha, noOfSeeds)
+% generic cross-validation function used to estimate the best parameter
+% (the one giving the best Test/Train error pair. Can only test 2
+% parameters at simultaneously at most, do NOT use vary more.
+% If no parameters to validate is given, then the function simply gives back
+% a cross-validation issued estimate of the test Error
+
+% split data in K fold (we will only create indices)
+
+% if lambdas is empty, then simply the code validates other parameters or
+% estimates test error
+noOfLambdas = max(1, length(lambdas));
+
+errorTr = zeros(noOfSeeds, K, noOfLambdas);
+errorTe = zeros(noOfSeeds, K, noOfLambdas);
+
+% K-fold cross validation
+for s = 1:noOfSeeds
+    
+    setSeed(s);
+    N = size(y,1);
+    idx = randperm(N);
+    Nk = floor(N/K);
+    for k = 1:K
+        idxCV(k,:) = idx(1+(k-1)*Nk:k*Nk);
+    end
+    
+    for i = 1:noOfLambdas
+        for k = 1:K
+            % get k'th subgroup in test, others in train
+            idxTe = idxCV(k,:);
+            idxTr = idxCV([1:k-1 k+1:end],:);
+            idxTr = idxTr(:);
+            yTe = y(idxTe);
+            tXTe = tX(idxTe,:);
+            yTr = y(idxTr);
+            tXTr = tX(idxTr,:);
+
+            fittingFunctionType = nargin(fittingFunction);
+
+            switch fittingFunctionType
+                case 2 % only direct LS uses 2 args
+                    beta = fittingFunction(yTr, tXTr);
+                case 3 % LS using GD, log. reg. and ride reg. use 3 args
+                    if ~isempty(alpha) % this means it's either LS GD or log. reg, same input args
+                        beta = fittingFunction(yTr, tXTr, alpha);
+                    else %otherwise it must be ridge regression
+                        beta = fittingFunction(yTr, tXTr, lambdas(i));
+                    end
+                case 4 % only penalized log. reg. used 4 arguments
+                    beta = fittingFunction(yTr, tXTr, alpha, lambdas(i));
+            end
+
+
+
+            % training and test cost
[…]

### 2014-10-31 youssef plays/youssefplaying.m

@@ -71,7 +71,7 @@ disp(' ');
 disp(['rank of data columns is :' num2str(ranktX)]);
 
 % toggling categorical variable encoding
-tX = tX(:,[1 (normInputVars+1)]);
+%tX = tX(:,[1 (normInputVars+1)]);
 
 % stupid mean fitting
 betaMean = mean(y_train);

### 2014-11-01 youssef plays/youssefplaying.m
added some comments + dummy variable testing on regression

@@ -33,12 +33,16 @@ noOfVariableValues = zeros(D, 1);
 % SUPPOSING categorial variables have AT MOST 15 classes
 maxClasses = 15;
 
+categorialVariableClasses = cell(81, D+1);
+
 for i=1:D
-    test = unique(X_train(:, i));
-    noOfVariableValues(i) = length(test);
+    temp = unique(X_train(:, i));
+    categorialVariableClasses{i} = temp;
+    noOfVariableValues(i) = length(temp);
 end
 
 catVars = find(noOfVariableValues < maxClasses);
+categorialVariableClasses = categorialVariableClasses(catVars);
 
 %% normalizing the features
 
@@ -60,12 +64,21 @@ y_train = (y_train-mean(y_train))/std(y_train);
 % Form (y,tX) to get regression data in matrix form
 tX = [ones(N,1) X_train(:, :)];
 
+%% replacing categorical variables with dummy ones
+
+dummyCategoricalVariables = dummyvar(X_train(:, catVars)+1);
+%txExt = [txExt(:, setdiff(1:(D+1), catVars)) dummyCategoricalVariables];
+
 %% correlation & normal distribution analysis
 
 % taking the 18th and 34th variables, and any polynomial transform of them,
 % and checking their pearson correlation with the output
 
-txSpecial = [myPoly(X_train(:, 18),5) myPoly(X_train(:, 34),10) ];
+% use the folowing line for testing dummy variables's correlation with the
+% output, comment out the next assignment to txSpecial
+txSpecial = dummyCategoricalVariables;
+
+%txSpecial = [myPoly(X_train(:, 18),5) myPoly(X_train(:, 34),10) ];
 statistics = [y_train txSpecial];
 % use this line if u want to plot the correlations for all the data instead
 %statistics = [y_train X_train];
@@ -76,24 +89,23 @@ statistics = [y_train txSpecial];
 
 normalityTests = circshift(normalityTests, -1);
 
-figure;
-spearmanGraph = abs(spearmanSig.*spearmanCoeffs);
-stem(spearmanGraph)
-title('correlated variables with coeff, 0 otherwise, spearman, , in absolute value')
+% figure;
+% spearmanGraph = abs(spearmanSig.*spearmanCoeffs);
+% stem(spearmanGraph)
+% title('correlated variables with coeff, 0 otherwise, spearman, , in absolute value')
 
 figure;
[…]

### 2014-11-01 youssef plays/youssefplaying.m
tested sqrt (see my comments & script)

@@ -160,6 +160,13 @@ meanErrTe = mean(rmseTe);
 disp(' ')
 disp(['estimated test error using transforms on 18th and 34th, gives test RMSE: ' num2str(meanErrTe)])
 
+txExt =[tX(:, setdiff(1:49, 35)) X_train(:, 18).^3 myPoly(X_train(:, 34), 6) (X_train(:, 34)-min(X_train(:, 34))).^0.5];
+[rmseTr, rmseTe] = genericKCV( y_train, txExt,...
+@leastSquares, @rmse, [], 4, [], 100);
+meanErrTe = mean(rmseTe);
+disp(' ')
+disp(['estimated test error using transforms on 18th and 34th + sqrt of 34th, gives test RMSE: ' num2str(meanErrTe)])
+
 %% adding good dummy variables
 clc
 

### 2014-11-02 decideRemove.m
added comments & some scripting

@@ -4,7 +4,7 @@ function [rA, idx] = decideRemove( A )
 % the ouput is zero and that all the row of v in A is 0 (except v corr v)
 % idx is the vector of deleted indx
 outputCulumn = size(A,2);
-maxCorr = max(A(:,outputCulumn));
+maxCorr = max(abs(A(:,outputCulumn)));
 threshold = 0.10*maxCorr;
 rA = A;
 idx =[];

### 2014-11-02 youssef plays/youssefplaying.m

@@ -4,8 +4,8 @@ clc
 clear all
 close all
 
-load('MexicoCity_regression.mat')
-%load('MexicoCity_classification.mat')
+%load('MexicoCity_regression.mat')
+load('MexicoCity_classification.mat')
 
 N = length(y_train); % data size
 D = size(X_train, 2); % dimensionality
@@ -59,7 +59,9 @@ stdX = std(X_train);
 X_train(:, normInputVars) = X_train(:, normInputVars)./repmat(stdX(:, normInputVars), N, 1);
 
 % normalizing the output variable, comment out for classification
-y_train = (y_train-mean(y_train))/std(y_train);
+%y_train = (y_train-mean(y_train))/std(y_train); % for regression
+%y_train = -y_train;
+y_train(y_train == -1) = 0; % for classification 
 
 % Form (y,tX) to get regression data in matrix form
 tX = [ones(N,1) X_train(:, :)];
@@ -76,10 +78,16 @@ dummyCategoricalVariables = dummyvar(X_train(:, catVars)+1);
 
 % use the folowing line for testing dummy variables's correlation with the
 % output, comment out the next assignment to txSpecial
-txSpecial = dummyCategoricalVariables;
+%txSpecial = dummyCategoricalVariables;
 
 %txSpecial = [myPoly(X_train(:, 18),5) myPoly(X_train(:, 34),10) ];
-statistics = [y_train txSpecial];
+%statistics = [y_train txSpecial];
+
+% for testing all polynomials of the interesting 11th and 24th
+% classification variables
+%statistics = [y_train  myPoly(X_train(:, 11), 5) myPoly(X_train(:, 24), 5)];
+statistics = [y_train  (X_train(:, 11)-min(X_train(:, 11))).^0.5 (X_train(:, 24)-min(X_train(:, 24))).^0.5];
+
 % use this line if u want to plot the correlations for all the data instead
 %statistics = [y_train X_train];
 
@@ -160,12 +168,12 @@ meanErrTe = mean(rmseTe);
 disp(' ')
 disp(['estimated test error using transforms on 18th and 34th, gives test RMSE: ' num2str(meanErrTe)])
 
[…]

### 2014-11-02 leastSquares.m
""

@@ -1,4 +1,5 @@
 function [ beta ] = leastSquares( y, Tx )
+% Compute beta using least squares
 
 beta = (Tx'*Tx)\(Tx'*y);
 

### 2014-11-02 loss01.m
""

@@ -1,6 +1,6 @@
 function loss = loss01( y,haty )
-%LOSS01 Summary of this function goes here
-%   Detailed explanation goes here
+%LOSS01 computes the 0-1 loss 
+
 N = length(y);
 loss = 1/N * sum(y~=haty);
 

### 2014-11-02 ridgeRegression.m
""

@@ -1,5 +1,5 @@
 function beta = ridgeRegression(y,tX,lambda)
-% compute beta using the ridgeRegression
+% compute beta using the ridgeRegression 
 [~,M] = size(tX);
 beta = (tX'*tX + lambda*eye(M))\tX'*y;
 

### 2014-11-02 analyzeKCVresults.m
commented code

@@ -3,6 +3,7 @@ function [bestParameter, achievedErrorTe ] = analyzeKCVresults( errorTe,...
 % plots the results from cross-validating the given parameter, for
 % inspecttion and also retrieves the optimal parameter value and returns it
 
+% averages the curves for different seeds into one
 avgErrorTe = mean(errorTe);
 avgErrorTr = mean(errorTr);
 
@@ -14,6 +15,8 @@ plot(parameterValues, avgErrorTe ,'r-','linewidth', 3); hold on
 plot(parameterValues, avgErrorTr,'b-','linewidth', 3); xlabel(parameterName);
 ylabel('error');
 
+% computes the best parameter (usually a penalizazion term lambda) that
+% achieves the best test Error, returns both
 [achievedErrorTe, bestParameter] = min(avgErrorTe);
 bestParameter = parameterValues(bestParameter);
 

### 2014-11-02 computeGradientLogReg.m

@@ -1,5 +1,6 @@
 function [ gradient ] = computeGradientLogReg( y, tX, beta )
-% computes the MSE gradient for the given data points
+% computes the gradient for logistic regression, using the given formulas
+% in the lectures
 
 gradient = tX'*(sigmoid(tX*beta) - y)/length(y);
 

### 2014-11-02 genericKCV.m

@@ -1,23 +1,28 @@
-function [errorTr, errorTe] = genericKCV( y, tX,...
+function [errorsTr, errorsTe] = genericKCV( y, tX,...
     fittingFunction, errorFunction, lambdas, K, alpha, noOfSeeds)
 % generic cross-validation function used to estimate the best parameter
-% (the one giving the best Test/Train error pair. Can only test 2
-% parameters at simultaneously at most, do NOT use vary more.
-% If no parameters to validate is given, then the function simply gives back
-% a cross-validation issued estimate of the test Error
-
-% split data in K fold (we will only create indices)
+% (the one giving the best Test/Train error pair (more emphasis on the
+% latter). It can be used with any fitting method (e.g. least squares) and
+% any error metric (e.g. rmse) to cross validate a parameter lambda (for ridgeregression for example)
+% or to simply estimate testError (in that case only a specific value of
+% lambda should be used, not multiple ones). To actually get the test error
+% at the end you simply need to take the mean of errorsTe in that case.
+% Alpha should also be given when the fitting function requires it, and the
+% noOfSeeds determines the number of test curves being averaged (more makes
+% the cross-validation more stable, but takes longer). K is the number of
+% folds
 
 % if lambdas is empty, then simply the code validates other parameters or
 % estimates test error
 noOfLambdas = max(1, length(lambdas));
 
-errorTr = zeros(noOfSeeds, K, noOfLambdas);
-errorTe = zeros(noOfSeeds, K, noOfLambdas);
+errorsTr = zeros(noOfSeeds, K, noOfLambdas);
+errorsTe = zeros(noOfSeeds, K, noOfLambdas);
 
 % K-fold cross validation
 for s = 1:noOfSeeds
     
+    % split data in K fold (we will only create indices)
     setSeed(s);
     N = size(y,1);
     idx = randperm(N);
@@ -52,23 +57,24 @@ for s = 1:noOfSeeds
                     beta = fittingFunction(yTr, tXTr, alpha, lambdas(i));
             end
 
-
-
             % training and test cost
-            errorTr(s, k, i) = errorFunction(yTr, tXTr*beta);
+            errorsTr(s, k, i) = errorFunction(yTr, tXTr*beta);
 
             % testing MSE using least squares
[…]

### 2014-11-02 gradientDescent.m

@@ -1,11 +1,16 @@
 function [ beta ] = gradientDescent(y, tX, maxIters, alpha, betaStart, lossFun, gradFun, lambda, verbose)
+% generic gradient descent function, it simply uses functions returning
+% the cost and its corresponding gradient, and lambda/alpha values if required, and the starting
+% point. Give 1 for the verbose value when you want to list the cost and
+% beta value at each iteration.
 
 beta = betaStart;
 
 minIters = maxIters/5;
 
+epsilon = 10^-2;
+
 for k = 1:maxIters
-    
 
     tempBeta = beta;
     tempBeta(1) = 0;
@@ -14,7 +19,6 @@ for k = 1:maxIters
     % computing the loss using the given function, plus norm squared of
     % beta * lambda as a regularization term
     L = lossFun( y, tX, beta ) + 1/2*1/length(y)*lambda*tempBetaNorm^2;
-    
     
     % comuting the gradient using the given function, if regularizing then
     % the regularization term's derivative is 2*lambda*beta
@@ -27,13 +31,13 @@ for k = 1:maxIters
     
     % tests whether the descent has converged ()
     
-    % when the norm of the beta changed by less than 1/1000, time to stop
-    if (max(abs(beta-oldBeta)./oldBeta) < eps) && (k > minIters)
+    % when the norm of the beta changed by less than eps, time to stop
+    if (max(abs(beta-oldBeta)./oldBeta) < epsilon) && (k > minIters)
         break
     end
 
     if verbose
-        % print current corst and beta
+        % print current cost and beta
         disp(['cost: ' num2str(L) ' and beta values ' num2str(beta(:)')]);
     end
 

### 2014-11-02 leastSquares.m

@@ -1,5 +1,5 @@
 function [ beta ] = leastSquares( y, Tx )
-% Compute beta using least squares
+% direct LS computation method, using the fomula from the lectures
 
 beta = (Tx'*Tx)\(Tx'*y);
 

### 2014-11-02 loss01.m

@@ -1,5 +1,5 @@
 function loss = loss01( y,haty )
-%LOSS01 computes the 0-1 loss 
+% implements the 0-1 loss error metric given in the project page
 
 N = length(y);
 loss = 1/N * sum(y~=haty);

### 2014-11-02 ridgeRegression.m

@@ -1,5 +1,7 @@
 function beta = ridgeRegression(y,tX,lambda)
-% compute beta using the ridgeRegression 
+% compute beta using the ridgeRegression, using the direct formula form the
+% lectures
+
 [~,M] = size(tX);
 beta = (tX'*tX + lambda*eye(M))\tX'*y;
 

### 2014-11-02 rmseClassification.m

@@ -1,4 +1,7 @@
 function  error  = rmseClassification( y, tXBeta )
+% computes the rmse between a given (categorical) output vector y and the
+% probalities obtained from logistic regression, as described on the
+% project page
 
 pHat = probEstimate(tXBeta);
 error = rmse(y, pHat);

### 2014-11-02 sigmoid.m

@@ -1,5 +1,5 @@
 function [ Y ] = sigmoid( X )
-
+% implements the sigmoid function used in logistic regression
 Y = 1./(1+exp(-X));
 
 end

### 2014-11-02 gradientDescent.m
fixed a small number making gradient descent use too many iterations (epsilon)

@@ -8,7 +8,7 @@ beta = betaStart;
 
 minIters = maxIters/5;
 
-epsilon = 10^-2;
+epsilon = 1/1000;
 
 for k = 1:maxIters
 

### 2014-11-02 analyzeKCVresults.m
last changes + comments

@@ -1,5 +1,5 @@
 function [bestParameter, achievedErrorTe ] = analyzeKCVresults( errorTe,...
-    errorTr, parameterName, parameterValues )
+    errorTr, parameterName, parameterValues, givenTitle, errorName )
 % plots the results from cross-validating the given parameter, for
 % inspecttion and also retrieves the optimal parameter value and returns it
 
@@ -12,7 +12,8 @@ plot(parameterValues, errorTe,'r-','color',[1 0.7 0.7]);
 hold on
 plot(parameterValues, errorTr,'b-','color',[0.7 0.7 1]);
 plot(parameterValues, avgErrorTe ,'r-','linewidth', 3); hold on
-plot(parameterValues, avgErrorTr,'b-','linewidth', 3); xlabel(parameterName);
+plot(parameterValues, avgErrorTr,'b-','linewidth', 3); xlabel(parameterName); ylabel(errorName)
+title(givenTitle);
 ylabel('error');
 
 % computes the best parameter (usually a penalizazion term lambda) that

### 2014-11-02 youssef plays/youssefplaying.m

@@ -275,50 +275,50 @@ for i = 1:noOfLambdas
     
 end
 
-% % takin all variables + sqrts of 11th and 24th
-txExt = [tX (X_train(:, 11)-min(X_train(:, 11))).^0.5 (X_train(:, 24)-min(X_train(:, 24))).^0.5];
-[rmseTr, rmseTe] = genericKCV( y_train, txExt,...
-@logisticRegression, @rmseClassification, [], 4, alpha, noOfSeeds);
-meanErrTe = mean(rmseTe);
-disp(' ')
-disp(['estimated test error using sqrts of 11th and 24th variables, gives test RMSE: ' num2str(meanErrTe)]);
-
-%taking some degrees (2nd)
-txExt = [tX  X_train(:, 11).^2 X_train(:, 24).^2];
-[rmseTr, rmseTe] = genericKCV( y_train, txExt,...
-@logisticRegression, @rmseClassification, [], 4, alpha, noOfSeeds);
-meanErrTe = mean(rmseTe);
-disp(' ')
-disp(['estimated test error using squares of 11th and 24th variables, gives test RMSE: ' num2str(meanErrTe)]);
+% % % takin all variables + sqrts of 11th and 24th
+% txExt = [tX (X_train(:, 11)-min(X_train(:, 11))).^0.5 (X_train(:, 24)-min(X_train(:, 24))).^0.5];
+% [rmseTr, rmseTe] = genericKCV( y_train, txExt,...
+% @logisticRegression, @rmseClassification, [], 4, alpha, noOfSeeds);
+% meanErrTe = mean(rmseTe);
+% disp(' ')
+% disp(['estimated test error using sqrts of 11th and 24th variables, gives test RMSE: ' num2str(meanErrTe)]);
+% 
+% %taking some degrees (2nd)
+% txExt = [tX  X_train(:, 11).^2 X_train(:, 24).^2];
+% [rmseTr, rmseTe] = genericKCV( y_train, txExt,...
+% @logisticRegression, @rmseClassification, [], 4, alpha, noOfSeeds);
+% meanErrTe = mean(rmseTe);
+% disp(' ')
+% disp(['estimated test error using squares of 11th and 24th variables, gives test RMSE: ' num2str(meanErrTe)]);
 
 % %taking sqrts + some degrees
[…]

### 2014-11-02 report/paper.tex
01 loss + decideRemove
report update -------------- draft (of final) version available at 20h today --------------------------- comments after 20h -------------------- modifications at 23h --------------------

@@
 \end{itemize}
+These functions are different machine learning methods.
 
+\textcolor{red}{Formula?}
+
+\section{Data observation}
+We have two data set. One for regression and one for classification. \begin{description}
+\item[Regression] Consists of output variables $\mathbf{y}$ and input variables $\mathbf{X}$.
+The number of example is $N = 1400$ and each input $\mathbb{x_n}$ has dimensionality $D = 48$.
+The first $34$ are real valued and the last $14$ are categorical.
+\item[Classification]
+
+\end{description}
+
+\section{Data visualization and cleaning}
+Most variables seem normally distributed with various means, but the variables 43-48 seem categorical (42 not really normal, looks like the distribution of $y$ actually, 2 Gaussian with different means)
+Kstest for normality only confirms that first 33 input variables are indeed random. something fishy about the variables 34-42 that SEEM normal (through histogram distribution)  but are not.\\
+
+The matrix $\tilde{X}$ has rank 36 for regression.
+That means 13 variables are not correlated to anything (given a threshold of 10 percent of the maximum correlation).
+Then $\tilde{X}$ is not ill conditioned.
 \section{Summary}

### 2014-11-02 report/paper.tex
report update - skeletton ------------------------ no time to continue before 17h comments.rtf will be used

--- 2014-10-28 report/paper.tex
@@
 \end{itemize}
+These functions are different machine learning methods.
 
+\textcolor{red}{Formula? Not even sure this section is useful}
+
+\section{Data observation}
+We have two data set. One for regression and one for classification. \begin{description}
+\item[Regression] Consists of output variables $\mathbf{y}$ and input variables $\mathbf{X}$.
+The number of example is $N = 1400$ and each input $\mathbf{x_n}$ has dimensionality $D = 48$.
+The first $34$ are real valued and the last $14$ are categorical.
+\item[Classification]
+
+\end{description}
+
+\section{Data visualization and cleaning}
+\textcolor{red}{Histogram, correlation, applied methods}
+
+\section{Best method : Ridge Regression}
+\textcolor{red}{explain what is the best method and why for our dataset, add some figures and results}
+\section{Feature transformations}
+\textcolor{red}{Different transformation (myPoly, sqrt, etc)}
 \section{Summary}
-In this report, we analyzed a regression dataset and found that ridge regression is a reasonable fit.
-We estimate that the average test error is 1.213 ($\pm$ 0.02).
-We tried some feature transformation and found that there is a small improvement giving us a test error of around 1.198 ($\pm$ 0.015).
-This improvement, however, is not significant.
+\textcolor{red}{summarize in a few lines and write down all the final results}
 

### 2014-11-02 report/paper.tex
up1

@@
 \item[Regression] Consists of output variables $\mathbf{y}$ and input variables $\mathbf{X}$.
-The number of example is $N = 1400$ and each input $\mathbf{x_n}$ has dimensionality $D = 48$.
-The first $34$ are real valued and the last $14$ are categorical.
-\item[Classification]
+The number of examples is $N = 1400$ and each input $\mathbf{x_n}$ has dimensionality $D = 48$.
+The first $34$ are real valued and the last $14$ are categorical, included $5$ that are binaries.
+\item[Classification] Consists of output variables $\mathbf{y}$ and input variables $\mathbf{X}$.
+The number of examples is $N = 1500$ and each input $\mathbf{x_n}$ has dimensionality $D = 35$.
+\textcolor{red}{check hist}
 
@@
 \section{Data visualization and cleaning}
-\textcolor{red}{Histogram, correlation, applied methods}
+\textcolor{red}{Histogram, correlation, applied methods}\\
+Most variables normally distributed with various means.
+The first 34 input variables are Gaussian and the last 14 are categorical.
+Given that, we normalized and centred the first 34 variables and let the others untouched.
+Plotting the histogram showed us that the data are compact and there is no outlier.
 

### 2014-11-02 analyzeKCVresults.m
some small changes

@@ -12,9 +12,9 @@ plot(parameterValues, errorTe,'r-','color',[1 0.7 0.7]);
 hold on
 plot(parameterValues, errorTr,'b-','color',[0.7 0.7 1]);
 plot(parameterValues, avgErrorTe ,'r-','linewidth', 3); hold on
-plot(parameterValues, avgErrorTr,'b-','linewidth', 3); xlabel(parameterName); ylabel(errorName)
+plot(parameterValues, avgErrorTr,'b-','linewidth', 3); xlabel(parameterName);
+ylabel(errorName)
 title(givenTitle);
-ylabel('error');
 
 % computes the best parameter (usually a penalizazion term lambda) that
 % achieves the best test Error, returns both

### 2014-11-02 report/paper.tex

@@
 Given that, we normalized and centred the first 34 variables and let the others untouched.
-Plotting the histogram showed us that the data are compact and there is no outlier.
+Plotting the histogram showed us that the data are compact and there is no outlier.\textcolor{red}{show figure hist?
+Or boxplot like in template}\\
+WHY WE ARE USING PEARSON:
+To help us with the task of finding which transforms could be useful in better predicting the output from the given predictors, we decided to use the help of the pearson linear correlarion coefficient.
+This coefficient basically gives a number indicating the strength (in its absolute value) of a linear model fit between the input and the data.
+Obviously, we could visually inspect multiple scatter plots between each variable and the output, and also with the other variables, but this can (for obvious reasons) be somewhat of an overkill and could lead to subjective mistakes.
+The pearson coefficient allows us to selectively choose wich variables (and which possible transforms accordingly) could be used to fit a good model, in a very fast manner (simply adding columns to tX and checking their corrrelation coefficients).
+Using this strategy, we could rapidly pinpoint which transforms were probably most useful.
+The spearman coefficient also performs a similar computation, but tests not only the linear relationship between variables, including any exponential relationship or such.
+Therefore for some transformations, even though a linearly-fit model would not find them useful, the spearman coefficient would rank them as correlated ( to the output and other variables) as the variable they are coming from.\\
 
+We checked the correlation between output and input variables.
+The correlation matrix helped us to discover which variables where correlated to no other variables nor the output.
+We could remove these variables.\textcolor{red}{More details to add here :youssef text}
+\section{Least Squares}
+Applying the least squares method on the data without cleaning yields a $RMSE$ of $0.48244$.
 \section{Best method : Ridge Regression}

### 2014-11-02 loss01Wrapper.m
small changes

@@ -1,0 +1,8 @@
+function [ error01 ] = loss01Wrapper(y, txBeta)
+% wrapper for 01 loss to have the same header as other error functions
+
+yhat = txBeta > 0;
+error01 = loss01( y, yhat );
+
+end
+

### 2014-11-02 report/paper.tex
up3

@@
 The correlation matrix helped us to discover which variables where correlated to no other variables nor the output.
-We could remove these variables.\textcolor{red}{More details to add here :youssef text}
+We could remove these variables.
 \section{Least Squares}
-Applying the least squares method on the data without cleaning yields a $RMSE$ of $0.48244$.
-\section{Best method : Ridge Regression}
-\textcolor{red}{explain what is the best method and why for our dataset, add some figures and results}
+The mean gives a $RMSE$ of 0.99964, over normalized $\mathbf{y}$.\\
+
+Applying the least squares method on the data without cleaning yields a $RMSE$ of $0.48244$. Given that, we see that the $RMSE$ is cut by half so the input data are meaningful.
+\section{Ridge Regression}
+\textcolor{red}{explain what is the best method and why for our dataset, add some figures and results}\\
+
+By visually inspecting the Pearson correlation graphs, a good strategy to spot well correlated input variables was to take the intersection of the two methods (we’re only considering ).
+To do that, we tried multiplying the two coefficients (for each variable) and taking the mean.
+Both strategies clearly showed the 18th and 34th variables are best correlated.
+We can limit our variable transform to these two (we only have 1400 data points, so with a dimension 48 we’re already lacking data, by introducing spurious transforms we would be adding fuel to the fire and causing overfitting.)
+\section{Logistic Regression}
+We tried to use simple logistic regression using all the variable, the estimated $RMSE$ test error was 0.27099.
+Also we tried using sqrt on the 11th and 24th variables which actually improves our test error to 0.26779.
+We can also use squares of the variables instead of the square root and applying this alone the test error decreased to 0.26715.
+Adding both transformations got us a decrease to a test error of 0.26474.
+We also tried to remove all categorical variables as we did in regression, but the test error increased to 0.3396, so their information was needed and cannot be ignored.
+
+We then found that, since Pearson correlation predicted better correlation of square roots than squares but squares were actually better able to decrease test error, it might be a good idea to include 3rd degree transforms of the 11th and 24th variables.
+Added them, the test error increased to 0.26656.
+Thus taking higher orders doesn’t seem to help.
+
+\textcolor{red}{to section transformation?}We then tried to include a dummy variable transform, but that increased the test error significantly to 0.32625, therefore we’re not going to use them.
 \section{Feature transformations}

### 2014-11-02 gradientDescent.m
decreased eps in gradient descent to allow for more iterations

@@ -8,7 +8,7 @@ beta = betaStart;
 
 minIters = maxIters/5;
 
-epsilon = 1/1000;
+epsilon = 1/10000;
 
 for k = 1:maxIters
 

### 2014-11-02 testCode.m

@@ -1,6 +1,8 @@
 % Run this to make sure your implementation is correct
 % An example of writing csv file is given in the end of this file
 clear all;
+close all
+clc
 
 % generating dummy dataset. regression
 beta = [1,2,3,4,5]';

### 2014-11-02 youssef plays/youssefplaying.m

@@ -4,8 +4,8 @@ clc
 clear all
 close all
 
-%load('MexicoCity_regression.mat')
-load('MexicoCity_classification.mat')
+load('MexicoCity_regression.mat')
+%load('MexicoCity_classification.mat')
 
 N = length(y_train); % data size
 D = size(X_train, 2); % dimensionality
@@ -59,9 +59,9 @@ stdX = std(X_train);
 X_train(:, normInputVars) = X_train(:, normInputVars)./repmat(stdX(:, normInputVars), N, 1);
 
 % normalizing the output variable, comment out for classification
-%y_train = (y_train-mean(y_train))/std(y_train); % for regression
+y_train = (y_train-mean(y_train))/std(y_train); % for regression
 %y_train = -y_train;
-y_train(y_train == -1) = 0; % for classification 
+%y_train(y_train == -1) = 0; % for classification 
 
 % Form (y,tX) to get regression data in matrix form
 tX = [ones(N,1) X_train(:, :)];

### 2014-11-02 report/paper.tex
up4

@@
 Or boxplot like in template}\\
+For regression, $\mathbf{\tilde{X}}$ has rank 49, so no redundancy in variables.
+For classification, the rank of $\mathbf{\tilde{X}}$ is 35 plus the one vector.
+this tells us that $\mathbf{\tilde{X}}$ is not ill-conditioned.\\
+We checked the correlation between output and input variables.
+The correlation matrix helped us to discover which variables where correlated to no other variables nor the output.
+We could remove these variables.
+We used the Pearson correlation.\\
+
 WHY WE ARE USING PEARSON:
@@
 
-We checked the correlation between output and input variables.
-The correlation matrix helped us to discover which variables where correlated to no other variables nor the output.
-We could remove these variables.
 \section{Least Squares}
@@
 Applying the least squares method on the data without cleaning yields a $RMSE$ of $0.48244$. Given that, we see that the $RMSE$ is cut by half so the input data are meaningful.
+\\
+
+We got the RMSE down to train error 0.3605 (with estimated test error 0.3805) using direct least Squares, simply by adding the squares of the 18th and 34th variables.
+When we added the squares of all the variables, we only got a train error of 0.3405 but this is clearly overfit because the corresponding test error estimate is 0.3765.\\
+
+We tested taking the squares of the 18th and the 34th variable individually and a curious thing happened: when we removed the 18th variable (squared) the test error increased to to 0.3881, but when we removed the 34th variable (squared) the test error increased much more significantly to 0.4987.
+
+\textcolor{red}{Move to transformation ?}By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the mypoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
+When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.48244
 \section{Ridge Regression}
@@
 \textcolor{red}{summarize in a few lines and write down all the final results}
+For the regression, we saw that the best method was Ridge Regression. The best parameter was $\lambda =  39.0694$\textcolor{red}{?} and we achieved a test error of 0.3073.\\
+
+For the classification, the best method was penalized logistic regression. We obtained different results depending the type of error we computed. \begin{description}
+\item[rmse] Best parameter is 0.0621 and we achieved a test error of 0.2663.
+\item[01 loss]Best parameter is 0.4642 and we achieved a test error of 0.0962.
+\item[log loss] Best parameter is 3.1623 and we achieved a test error of 0.2812.
+\end{description}
 

### 2014-11-02 report/paper.tex
up5 -> need to add figures now

@@
 
+\textcolor{red}{need to move that somewhere}Note that all the gradient descent methods used the eps threshold (machine dependant) in Matlab in the gradient descent function, to allow for a maximum number of iterations.
+This might make the gradient descent process slow though, so feel free to set another number if needed.
+We changed it later to 1/1000, so the numbers for the test Errors might be different depending on this parameter when you run the code (the global trends are the same though), since the number of iterations will be different.
+
 \section{Data visualization and cleaning}
@@
 \section{Feature transformations}
-\textcolor{red}{Different transformation (myPoly, sqrt, etc)}
+\textcolor{red}{Different transformation (myPoly, sqrt, dummyvar,etc)}\\
+
+We tried dummy variables for all regression categorical ones.
+We could only find a significant linear correlation (using Spearman) for the 2nd class of the variable 44, 2nd and 4th class of the 47th.\\
+
+We tried keeping the polynomial transforms and removed all categorical variables. The test error went down to 0.33043.\\
+
+Adding back the mentioned dummy variables 44 and 47 actually increased the test error to 0.33105. This lead the matrix $\mathbf{\tilde{X}}$ to become ill conditioned.\\
+
+We also tried adding square root of the 18th and 34th variable, that means $\sqrt{|X_{n18}|}$ and $\sqrt{|X_{n34}|}$ of $\mathbf{X}$.
+With the 18th the test error went up to 0.32137 and with the 34th it went up to 0.39349.
+So this doesn't help and we decided not tu use square root transformation.\\
+
+We decided that we’re going to ignore all categorical variables for regression, and also all the un-correlated variables, giving a test error of 0.33173 while including the significant transformations of the 18th and 34th variables discussed before.
+We decided to opt for a simpler model because of Occam’s Razor.
 \section{Summary}

### 2014-11-02 leastSquaresGD.m
small bug fixes

@@ -1,12 +1,12 @@
 function beta = leastSquaresGD(y,tX,alpha)
 % compute beta using the gradient descent with least squares
 
-maxIters = 2000;
+maxIters = 5000;
 lambda = 0;
 betaStart = zeros(size(tX,2),1);
 
 beta = gradientDescent(y,tX, maxIters, alpha, betaStart, @computeCostLS, ...
-    @computeGradientLS,lambda, 1);
+    @computeGradientLS,lambda, 0);
 
 end
 

### 2014-11-02 logLoss.m

@@ -17,6 +17,8 @@ p = p(:);
 N = length(y);
 
 % compute the logLoss
-lo = -(1/N).*sum((y.*log(p)+(1-p).*log(1-p)));
+% sometimes p = 1, then the log is not defined. So we ignore those points
+% by using the nanmean function, which doesn't take NaNs into account
+lo = -nanmean((y.*log(p)+(1-y).*log(1-p)));
 
 

### 2014-11-02 penLogisticRegression.m

@@ -2,7 +2,7 @@ function beta = penLogisticRegression(y,tX,alpha,lambda)
 % computes the logistic regression beta result using gradient descent.
 
 % 2000 iterations should be enough
-maxIters = 2000;
+maxIters = 5000;
 
 % starting from the beta = D-dimensional zero vector
 betaStart = zeros(size(tX, 2), 1);

### 2014-11-02 youssef plays/youssefplaying.m

@@ -4,8 +4,8 @@ clc
 clear all
 close all
 
-load('MexicoCity_regression.mat')
-%load('MexicoCity_classification.mat')
+%load('MexicoCity_regression.mat')
+load('MexicoCity_classification.mat')
 
 N = length(y_train); % data size
 D = size(X_train, 2); % dimensionality
@@ -59,9 +59,9 @@ stdX = std(X_train);
 X_train(:, normInputVars) = X_train(:, normInputVars)./repmat(stdX(:, normInputVars), N, 1);
 
 % normalizing the output variable, comment out for classification
-y_train = (y_train-mean(y_train))/std(y_train); % for regression
+%y_train = (y_train-mean(y_train))/std(y_train); % for regression
 %y_train = -y_train;
-%y_train(y_train == -1) = 0; % for classification 
+y_train(y_train == -1) = 0; % for classification 
 
 % Form (y,tX) to get regression data in matrix form
 tX = [ones(N,1) X_train(:, :)];
@@ -265,15 +265,15 @@ noOfLambdas = 10;
 RRrmses = zeros(1, noOfLambdas);
 lambdas = logspace(-2, 2, noOfLambdas);
 
-disp(' ');
-for i = 1:noOfLambdas
-
-    betaPLR = penLogisticRegression(y_train, tX, alpha, lambdas(i));
-    temp = tX*betaPLR;
-    RRrmses(i) = rmseClassification(y_train, temp);
-    disp(['for lambda = ' num2str(lambdas(i)) ' penalized logistic regression fits with training RMSE ' num2str(RRrmses(i))])
-    
-end
+% disp(' ');
+% for i = 1:noOfLambdas
+% 
+%     betaPLR = penLogisticRegression(y_train, tX, alpha, lambdas(i));
+%     temp = tX*betaPLR;
+%     RRrmses(i) = rmseClassification(y_train, temp);
+%     disp(['for lambda = ' num2str(lambdas(i)) ' penalized logistic regression fits with training RMSE ' num2str(RRrmses(i))])
+%     
+% end
 
 % % % takin all variables + sqrts of 11th and 24th
 % txExt = [tX (X_train(:, 11)-min(X_train(:, 11))).^0.5 (X_train(:, 24)-min(X_train(:, 24))).^0.5];
[…]

### 2014-11-02 report/paper.tex
report update -> (draft) final version ----------- need to check one last time if everything ok + remove red comments

@@
 \begin{abstract}
-In this report, we discuss our implementation and findings for the project-I. \textcolor{red}{Todo at the end}
+In this report, we discuss our implementation and findings for the project-I.
+We fit several methods to the regression dataset and compared them.
+We found that the input matrix is not ill conditioned and that ridge regression gives a better test error.
+Trying non-linear model doesn't help so the linear model is appropriate for this data set.
+We investigated some feature transformation.
+We also used a classification data set for which we found that the input matrix has full rank which means all inputs are correlated.
+We saw that the best test error can be achieved by computing the log loss.
 
@@
 \section{Data observation}
+\label{sec:datadescr}
 We have two data set. One for regression and one for classification. \begin{description}
@@
 
-\section{Data visualization and cleaning}
-\textcolor{red}{Histogram, correlation, applied methods}\\
+\section{Regression}
+\subsection{Data visualization and cleaning}
 Most variables normally distributed with various means.
 The first 34 input variables are Gaussian and the last 14 are categorical.
+The boxplot in Figure \ref{fig:boxplot} shows us that the data is not centred.
 Given that, we normalized and centred the first 34 variables and let the others untouched.
-Plotting the histogram showed us that the data are compact and there is no outlier.\textcolor{red}{show figure hist?
-Or boxplot like in template}\\
-For regression, $\mathbf{\tilde{X}}$ has rank 49, so no redundancy in variables.
-For classification, the rank of $\mathbf{\tilde{X}}$ is 35 plus the one vector.
-this tells us that $\mathbf{\tilde{X}}$ is not ill-conditioned.\\
+Plotting the histogram as shown if Figure \ref{fig:outputhist} showed us that the data are compact and there is no outlier, but we can see a second bump.\\
+For regression, $\mathbf{\tilde{X}}$ has rank 49, included the one vector, so no redundancy in variables.\\
+
+To help us with the task of finding which transforms could be useful in better predicting the output from the given predictors, we decided to use the help of the pearson linear correlarion coefficient.
+This coefficient basically gives a number indicating the strength (in its absolute value) of a linear model fit between two variables.
+Obviously, we could visually inspect multiple scatter plots between each variable and the output, and also with the other variables, but this can (for obvious reasons) be somewhat of an overkill and could lead to subjective mistakes.
+The pearson coefficient allows us to selectively choose wich variables (and which possible transforms accordingly) could be used to fit a good model, in a very fast manner (simply adding columns to tX and checking their corrrelation coefficients).
+The spearman coefficient could also be used, but is much less adapted for linear fitting in our case.\\
+
 We checked the correlation between output and input variables.
@@
 We could remove these variables.
-We used the Pearson correlation.\\
+We used the Pearson correlation.
+The variables that are neither significantly correlated with the output, nor with the other input variables, were found and were later removed from our model for simplicity and to improve generalization performance
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{images/boxplot_xtrain_reg.eps}
+\caption{Boxplot of all input variables}
+\label{fig:boxplot}
+\end{figure}
 
-WHY WE ARE USING PEARSON:
-To help us with the task of finding which transforms could be useful in better predicting the output from the given predictors, we decided to use the help of the pearson linear correlarion coefficient.
-This coefficient basically gives a number indicating the strength (in its absolute value) of a linear model fit between the input and the data.
-Obviously, we could visually inspect multiple scatter plots between each variable and the output, and also with the other variables, but this can (for obvious reasons) be somewhat of an overkill and could lead to subjective mistakes.
-The pearson coefficient allows us to selectively choose wich variables (and which possible transforms accordingly) could be used to fit a good model, in a very fast manner (simply adding columns to tX and checking their corrrelation coefficients).
-Using this strategy, we could rapidly pinpoint which transforms were probably most useful.
-The spearman coefficient also performs a similar computation, but tests not only the linear relationship between variables, including any exponential relationship or such.
-Therefore for some transformations, even though a linearly-fit model would not find them useful, the spearman coefficient would rank them as correlated ( to the output and other variables) as the variable they are coming from.\\
+\begin{figure}
+\centering
+\includegraphics[width=0.45\textwidth]{images/hist_output.eps}
+\caption{Histogram of output variable $\mathbf{y}$}
+\label{fig:outputhist}
+\end{figure}
 
-\section{Least Squares}
-The mean gives a $RMSE$ of 0.99964, over normalized $\mathbf{y}$.\\
+\subsection{Least Squares}
 
-Applying the least squares method on the data without cleaning yields a $RMSE$ of $0.48244$. Given that, we see that the $RMSE$ is cut by half so the input data are meaningful.
+The mean gives a RMSE of 0.99964, over normalized $\mathbf{y}$.\\
+
+Applying the least squares method on the data without cleaning yields a training RMSE of $0.48244$. Given that, we see that the RMSE is cut by half so the input data are meaningful.
 \\
@@
 
-\textcolor{red}{Move to transformation ?}By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the mypoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
-When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.48244
-\section{Ridge Regression}
-\textcolor{red}{explain what is the best method and why for our dataset, add some figures and results}\\
-
+\subsection{Ridge Regression}
+\label{sec:regridgreg}
 By visually inspecting the Pearson correlation graphs, a good strategy to spot well correlated input variables was to take the intersection of the two methods (we’re only considering ).
@@
 Both strategies clearly showed the 18th and 34th variables are best correlated.
-We can limit our variable transform to these two (we only have 1400 data points, so with a dimension 48 we’re already lacking data, by introducing spurious transforms we would be adding fuel to the fire and causing overfitting.)
-\section{Logistic Regression}
-We tried to use simple logistic regression using all the variable, the estimated $RMSE$ test error was 0.27099.
+We can limit our variable transform to these two (we only have 1400 data points, so with a dimension 48 we’re already lacking data, by introducing spurious transforms we would be adding fuel to the fire and causing overfitting.)\\
+We have also done a cross validation using seeds. The result is shown in Figure \ref{fig:ridreggcv}.
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{images/ridgeRegCV.eps}
+\caption{Cross validation : Ridge regression with different lambda values }
+\label{fig:ridreggcv}
+\end{figure}
+
+\subsection{Logistic Regression}
+We tried to use simple logistic regression using all the variable, the estimated RMSE test error was 0.27099.
 Also we tried using sqrt on the 11th and 24th variables which actually improves our test error to 0.26779.
@@
 We also tried to remove all categorical variables as we did in regression, but the test error increased to 0.3396, so their information was needed and cannot be ignored.
+\subsection{Feature transformations}
+By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the mypoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
+When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.3805.\\
 
-We then found that, since Pearson correlation predicted better correlation of square roots than squares but squares were actually better able to decrease test error, it might be a good idea to include 3rd degree transforms of the 11th and 24th variables.
-Added them, the test error increased to 0.26656.
+We then found that, since Pearson correlation predicted better correlation of square roots than squares but squares were actually better able to decrease test error, it might be a good idea to include 3rd degree transforms of the 11th and 24th variables for logistic regression.
+After having added them, the test error increased to 0.26656.
 Thus taking higher orders doesn’t seem to help.
+We then tried to include a dummy variable transform, but that increased the test error significantly to 0.32625, therefore we’re not going to use them.
 
-\textcolor{red}{to section transformation?}We then tried to include a dummy variable transform, but that increased the test error significantly to 0.32625, therefore we’re not going to use them.
-\section{Feature transformations}
-\textcolor{red}{Different transformation (myPoly, sqrt, dummyvar,etc)}\\
+\section{Classification}
+
+\subsection{visualization and cleaning}
+
+For classification, the rank of $\mathbf{\tilde{X}}$ is 35 plus the one vector.
+Given the data description from section \ref{sec:datadescr}, this tells us that $\mathbf{\tilde{X}}$ is not ill-conditioned.\\
+
+\subsection{Ridge Regression}
+For the classification, we repeated the procedure describe in section \ref{sec:regridgreg}.
+We found that the 11th and the 24th variables are best correlated with the output, therefore we used feature transforms on those.
+\textcolor{red}{what transformation?
+sqrt etc but only on these variables?
+}
+
+\subsection{Penalized logistic regression}
+We could improve the test error by cross-validation.
+The Figure \ref{fig:crossval} shows the three different test error RMSE, 0-1 loss and logloss.
+We used different numbers of seed to have a better prediction for our test error.
+\begin{figure}
+\centering
+\subfigure[0-1 loss]{\includegraphics[width=0.5\textwidth]{images/penLogRegCrossVall01Loss.eps}}
+\subfigure[rmse]{\includegraphics[width=0.5\textwidth]{images/penLogRegrmse.eps}}
+\subfigure[log loss]{\includegraphics[width=0.5\textwidth]{images/penLogRegLogloss.eps}}
+\caption{Cross-validation : penalized logistic regression with different lambda values}
+\label{fig:crossval}
+\end{figure} 
+\subsection{Feature transformations}
 
@@
 
-We decided that we’re going to ignore all categorical variables for regression, and also all the un-correlated variables, giving a test error of 0.33173 while including the significant transformations of the 18th and 34th variables discussed before.
+We decided to ignore all the un-correlated variables for regression, giving a test error of 0.33173 while including the significant transformations of the 18th and 34th variables discussed before.
 We decided to opt for a simpler model because of Occam’s Razor.
+
 \section{Summary}
-\textcolor{red}{summarize in a few lines and write down all the final results}
-For the regression, we saw that the best method was Ridge Regression. The best parameter was $\lambda =  39.0694$\textcolor{red}{?} and we achieved a test error of 0.3073.\\
+As we have seen, the cross validation helps to have a better test error.\\
+
+We obtained these RMSE for regression.
+\begin{description}
+\item[leastSquaresGD] We achieved a RMSE of 0.6852
+\item[leastSquares] We achieved a RMSE of 0.3317
+\item[RidgeRegression] We achieved a RMSE of 0.3073
+\end{description}
+For the regression, we saw that the best method was Ridge Regression. The best parameter was $\lambda =  39.0694$\textcolor{red}{?} and we achieved a test error of 0.3073.
 
 For the classification, the best method was penalized logistic regression. We obtained different results depending the type of error we computed. \begin{description}
-\item[rmse] Best parameter is 0.0621 and we achieved a test error of 0.2663.
-\item[01 loss]Best parameter is 0.4642 and we achieved a test error of 0.0962.
-\item[log loss] Best parameter is 3.1623 and we achieved a test error of 0.2812.
+\item[RMSE] We achieved a test error of 0.2644.
+\item[01 loss]We achieved a test error of 0.09486.
+\item[log loss] We achieved a test error of 0.25149.
 \end{description}
+With logistic regression we obtained different results depending the type of error we computed. \begin{description}
+\item[rmse] We achieved a test error of 0.2644.
+\item[01 loss]We achieved a test error of 0.09486.
+\item[log loss] We achieved a test error of 0.2637.
+\end{description}
+We see that there is only a small improvement with the log loss error.
 

### 2014-11-03 report/paper.tex

@@
 \centering
-\subfigure[0-1 loss]{\includegraphics[width=0.5\textwidth]{images/penLogRegCrossVall01Loss.eps}}
-\subfigure[rmse]{\includegraphics[width=0.5\textwidth]{images/penLogRegrmse.eps}}
-\subfigure[log loss]{\includegraphics[width=0.5\textwidth]{images/penLogRegLogloss.eps}}
+\subfigure[0-1 loss]{\includegraphics[width=1\textwidth]{images/penLogRegCrossVall01Loss.eps}}
+\subfigure[rmse]{\includegraphics[width=1\textwidth]{images/penLogRegrmse.eps}}
+\subfigure[log loss]{\includegraphics[width=1\textwidth]{images/penLogRegLogloss.eps}}
 \caption{Cross-validation : penalized logistic regression with different lambda values}

### 2014-11-03 report/paper.tex
update

@@
 
-\textcolor{red}{Formula? Not even sure this section is useful}
-
 \section{Data observation}
@@
 The first $34$ are real valued and the last $14$ are categorical, included $5$ that are binaries.
-\item[Classification] Consists of output variables $\mathbf{y}$ and input variables $\mathbf{X}$.
-The number of examples is $N = 1500$ and each input $\mathbf{x_n}$ has dimensionality $D = 35$.
-\textcolor{red}{check hist}
+\item[Classification] Consists of output variables $\mathbf{y}$ and input variables $\mathbf{X}$. The number of examples is $N = 1500$ and each input $\mathbf{x_n}$ has dimensionality $D = 35$.
 
@@
 
-\textcolor{red}{need to move that somewhere}Note that all the gradient descent methods used the eps threshold (machine dependant) in Matlab in the gradient descent function, to allow for a maximum number of iterations.
+Note that all the gradient descent methods used the eps threshold (machine dependant) in Matlab in the gradient descent function, to allow for a maximum number of iterations.
 This might make the gradient descent process slow though, so feel free to set another number if needed.
@@
 
-We tested taking the squares of the 18th and the 34th variable individually and a curious thing happened: when we removed the 18th variable (squared) the test error increased to to 0.3881, but when we removed the 34th variable (squared) the test error increased much more significantly to 0.4987.
+We tested taking the squares of the 18th and the 34th variable individually and a curious thing happened: when we removed the 18th variable (squared) the test error increased to to 0.3881, but when we removed the 34th variable (squared) the test error increased much more significantly to 0.4987.\\
 
-\subsection{Ridge Regression}
 \label{sec:regridgreg}
@@
 We can limit our variable transform to these two (we only have 1400 data points, so with a dimension 48 we’re already lacking data, by introducing spurious transforms we would be adding fuel to the fire and causing overfitting.)\\
-We have also done a cross validation using seeds. The result is shown in Figure \ref{fig:ridreggcv}.
+
+\subsection{Ridge Regression}
+After applying least squares, we have also done a cross validation using seeds. The result is shown in Figure \ref{fig:ridreggcv}.
 \begin{figure}
@@
 \end{figure}
+We realized that the RMSE decreased for a parameter $\lambda = 39.0694$ to a RMSE of $0.3073$. We can see that on the Figure \ref{fig:ridreggcv}.
 
-\subsection{Logistic Regression}
-We tried to use simple logistic regression using all the variable, the estimated RMSE test error was 0.27099.
-Also we tried using sqrt on the 11th and 24th variables which actually improves our test error to 0.26779.
-We can also use squares of the variables instead of the square root and applying this alone the test error decreased to 0.26715.
-Adding both transformations got us a decrease to a test error of 0.26474.
-We also tried to remove all categorical variables as we did in regression, but the test error increased to 0.3396, so their information was needed and cannot be ignored.
 \subsection{Feature transformations}
-By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the mypoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
-When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.3805.\\
-
-We then found that, since Pearson correlation predicted better correlation of square roots than squares but squares were actually better able to decrease test error, it might be a good idea to include 3rd degree transforms of the 11th and 24th variables for logistic regression.
-After having added them, the test error increased to 0.26656.
-Thus taking higher orders doesn’t seem to help.
-We then tried to include a dummy variable transform, but that increased the test error significantly to 0.32625, therefore we’re not going to use them.
-
-\section{Classification}
-
-\subsection{visualization and cleaning}
-
-For classification, the rank of $\mathbf{\tilde{X}}$ is 35 plus the one vector.
-Given the data description from section \ref{sec:datadescr}, this tells us that $\mathbf{\tilde{X}}$ is not ill-conditioned.\\
-
-\subsection{Ridge Regression}
-For the classification, we repeated the procedure describe in section \ref{sec:regridgreg}.
-We found that the 11th and the 24th variables are best correlated with the output, therefore we used feature transforms on those.
-\textcolor{red}{what transformation?
-sqrt etc but only on these variables?
-}
-
-\subsection{Penalized logistic regression}
-We could improve the test error by cross-validation.
-The Figure \ref{fig:crossval} shows the three different test error RMSE, 0-1 loss and logloss.
-We used different numbers of seed to have a better prediction for our test error.
-\begin{figure}
-\centering
-\subfigure[0-1 loss]{\includegraphics[width=1\textwidth]{images/penLogRegCrossVall01Loss.eps}}
-\subfigure[rmse]{\includegraphics[width=1\textwidth]{images/penLogRegrmse.eps}}
-\subfigure[log loss]{\includegraphics[width=1\textwidth]{images/penLogRegLogloss.eps}}
-\caption{Cross-validation : penalized logistic regression with different lambda values}
-\label{fig:crossval}
-\end{figure} 
-\subsection{Feature transformations}
-
 We tried dummy variables for all regression categorical ones.
@@
 
+By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the mypoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
+When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.3805.\\
+
+\section{Classification}
+
+\subsection{visualization and cleaning}
+
+For classification, the rank of $\mathbf{\tilde{X}}$ is 35 plus the one vector.
+Given the data description from section \ref{sec:datadescr}, this tells us that $\mathbf{\tilde{X}}$ is not ill-conditioned.\\ The range of $\mathbf{y}$ is between $-1$ and $1$.
+We also modified the values $-1$ from $\mathbf{y}$ to 0 to be consistent with the book.
+
+\subsection{Logistic Regression}
+For the classification, we repeated the procedure describe in section \ref{sec:regridgreg}.
+We found that the 11th and the 24th variables are best correlated with the output, therefore we used feature transforms on those as described in section \ref{sec:ftransclass}.
+
+\subsection{Feature transformations}
+\label{sec:ftransclass}
+We tried to use simple logistic regression using all the variable, the estimated RMSE test error was 0.27099.
+Also we tried using sqrt on the 11th and 24th variables which actually improves our test error to 0.26779.
+We can also use squares of the variables instead of the square root and applying this alone the test error decreased to 0.26715.
+Adding both transformations got us a decrease to a test error of 0.26474.
+We also tried to remove all categorical variables as we did in regression, but the test error increased to 0.3396, so their information was needed and cannot be ignored.\\
+
+We then found that, since Pearson correlation predicted better correlation of square roots than squares but squares were actually better able to decrease test error, it might be a good idea to include 3rd degree transforms of the 11th and 24th variables for logistic regression.
+After having added them, the test error increased to 0.26656.
+Thus taking higher orders doesn’t seem to help.
+We then tried to include a dummy variable transform, but that increased the test error significantly to 0.32625, therefore we’re not going to use them.
+
+\subsection{Penalized logistic regression}
+We could improve the test error by cross-validation.
+The Figure \ref{fig:crossval} shows the three different test error RMSE, 0-1 loss and logloss.
+We used different numbers of seed to have a better prediction for our test error.
+\begin{figure}
+\centering
+\subfigure[0-1 loss]{\includegraphics[width=1\textwidth]{images/penLogRegCrossVall01Loss.eps}}
+\subfigure[rmse]{\includegraphics[width=1\textwidth]{images/penLogRegrmse.eps}}
+\subfigure[log loss]{\includegraphics[width=1\textwidth]{images/penLogRegLogloss.eps}}
+\caption{Cross-validation : penalized logistic regression with different lambda values}
+\label{fig:crossval}
+\end{figure} 
 \section{Summary}
@@
 \end{description}
-For the regression, we saw that the best method was Ridge Regression. The best parameter was $\lambda =  39.0694$\textcolor{red}{?} and we achieved a test error of 0.3073.
+For the regression, we saw that the best method was Ridge Regression. The best parameter was $\lambda =  39.0694$ and we achieved a test error of 0.3073.
 
@@
 \end{description}
-We see that there is only a small improvement with the log loss error.
+We see that there is only a small improvement with the log loss error.\\
+The best error we achieved is through the logistic regression method and using the 0-1 loss error.
 

### 2014-11-03 report/paper.tex
update

@@
 \centering
-\includegraphics[width=1\textwidth]{images/boxplot_xtrain_reg.eps}
-\caption{Boxplot of all input variables}
+\includegraphics[width=0.7\textwidth]{images/boxplot_xtrain_reg.eps}
+\caption{Boxplot of all input variables for regression (except categorical)}
 \label{fig:boxplot}
@@
 Both strategies clearly showed the 18th and 34th variables are best correlated.
-We can limit our variable transform to these two (we only have 1400 data points, so with a dimension 48 we’re already lacking data, by introducing spurious transforms we would be adding fuel to the fire and causing overfitting.)\\
+We can limit our variable transform to these two (we only have 1400 data points, so with a dimension 48 we’re already lacking data, by introducing spurious transforms we would be adding fuel to the fire and causing overfitting.)
 
@@
 \centering
-\includegraphics[width=1\textwidth]{images/ridgeRegCV.eps}
+\includegraphics[width=0.5\textwidth]{images/ridgeRegCV.jpg}
 \caption{Cross validation : Ridge regression with different lambda values }
@@
 By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the mypoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
-When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.3805.\\
-
+When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.3805.
 \section{Classification}
-
 \subsection{visualization and cleaning}
@@
 We also modified the values $-1$ from $\mathbf{y}$ to 0 to be consistent with the book.
-
+We show the boxplot for $\mathbf{X}$ in Figure \ref{fig:boxplotclass}.
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{images/boxplot_class.eps}
+\caption{Boxplot for input variable $\mathbf{X}$ for classification}
+\label{fig:boxplotclass}
+\end{figure}
 \subsection{Logistic Regression}
@@
 We found that the 11th and the 24th variables are best correlated with the output, therefore we used feature transforms on those as described in section \ref{sec:ftransclass}.
-
 \subsection{Feature transformations}
@@
 We then tried to include a dummy variable transform, but that increased the test error significantly to 0.32625, therefore we’re not going to use them.
-
 \subsection{Penalized logistic regression}
@@
 \centering
-\subfigure[0-1 loss]{\includegraphics[width=1\textwidth]{images/penLogRegCrossVall01Loss.eps}}
-\subfigure[rmse]{\includegraphics[width=1\textwidth]{images/penLogRegrmse.eps}}
-\subfigure[log loss]{\includegraphics[width=1\textwidth]{images/penLogRegLogloss.eps}}
+\subfigure[0-1 loss]{\includegraphics[width=0.6\textwidth]{images/penLogRegCrossVall01Loss.jpg}}
+\subfigure[rmse]{\includegraphics[width=0.6\textwidth]{images/penLogRegrmse.jpg}}
+\subfigure[log loss]{\includegraphics[width=0.6\textwidth]{images/penLogRegLogloss.jpg}}
 \caption{Cross-validation : penalized logistic regression with different lambda values}
@@
 \subsubsection*{Acknowledgments}
+We would like to thanks the teaching team for their availability during the exercise session but also on the forum. We also want to thanks Youssef who organized the milestone of the project.
 

### 2014-11-03 report/paper.tex
final update

@@
 Trying non-linear model doesn't help so the linear model is appropriate for this data set.
-We investigated some feature transformation.
+We investigated some feature transformations.
 We also used a classification data set for which we found that the input matrix has full rank which means all inputs are correlated.
-We saw that the best test error can be achieved by computing the log loss.
+We saw that the best test error can be achieved by computing the 0-1 loss using logistic regression.
 
@@
 
+\section{Regression}
 Note that all the gradient descent methods used the eps threshold (machine dependant) in Matlab in the gradient descent function, to allow for a maximum number of iterations.
@@
 We changed it later to 1/1000, so the numbers for the test Errors might be different depending on this parameter when you run the code (the global trends are the same though), since the number of iterations will be different.
-
-\section{Regression}
 \subsection{Data visualization and cleaning}
-Most variables normally distributed with various means.
+Most variables are normally distributed with various means.
 The first 34 input variables are Gaussian and the last 14 are categorical.
@@
 Given that, we normalized and centred the first 34 variables and let the others untouched.
-Plotting the histogram as shown if Figure \ref{fig:outputhist} showed us that the data are compact and there is no outlier, but we can see a second bump.\\
+Plotting the histogram of $\mathbf{y}$ as shown if Figure \ref{fig:outputhist} showed us that the data is compact and there is no outlier, but we can see a second bump.\\
 For regression, $\mathbf{\tilde{X}}$ has rank 49, included the one vector, so no redundancy in variables.\\
 
-To help us with the task of finding which transforms could be useful in better predicting the output from the given predictors, we decided to use the help of the pearson linear correlarion coefficient.
+To help us with the task of finding which transforms could be useful in better predicting the output from the given predictors, we decided to use the help of the Pearson linear correlation coefficient.
 This coefficient basically gives a number indicating the strength (in its absolute value) of a linear model fit between two variables.
 Obviously, we could visually inspect multiple scatter plots between each variable and the output, and also with the other variables, but this can (for obvious reasons) be somewhat of an overkill and could lead to subjective mistakes.
-The pearson coefficient allows us to selectively choose wich variables (and which possible transforms accordingly) could be used to fit a good model, in a very fast manner (simply adding columns to tX and checking their corrrelation coefficients).
-The spearman coefficient could also be used, but is much less adapted for linear fitting in our case.\\
+The Pearson coefficient allows us to selectively choose which variables (and which possible transforms accordingly) could be used to fit a good model, in a very fast manner (simply adding columns to $\mathbf{\tilde{X}}$ and checking their correlation coefficients).
+The Spearman coefficient could also be used, but is much less adapted for linear fitting in our case.\\
 
@@
 We used the Pearson correlation.
-The variables that are neither significantly correlated with the output, nor with the other input variables, were found and were later removed from our model for simplicity and to improve generalization performance
+The variables that are neither significantly correlated with the output, nor with the other input variables, were found and were later removed from our model for simplicity and to improve generalization performance.
 \begin{figure}
@@
 
-The mean gives a RMSE of 0.99964, over normalized $\mathbf{y}$.\\
+The mean gives a RMSE of 0.99964, over normalized $\mathbf{y}$.
 
 Applying the least squares method on the data without cleaning yields a training RMSE of $0.48244$. Given that, we see that the RMSE is cut by half so the input data are meaningful.
-\\
 
-We got the RMSE down to train error 0.3605 (with estimated test error 0.3805) using direct least Squares, simply by adding the squares of the 18th and 34th variables.
-When we added the squares of all the variables, we only got a train error of 0.3405 but this is clearly overfit because the corresponding test error estimate is 0.3765.\\
+We got the RMSE down to train error 0.3605 (with estimated test error 0.3805) using direct least squares, simply by adding the squares of the 18th and 34th variables.
+When we added the squares of all the variables, we only got a train error of 0.3405 but this is clearly overfit because the corresponding test error estimate is 0.3765.
 
-We tested taking the squares of the 18th and the 34th variable individually and a curious thing happened: when we removed the 18th variable (squared) the test error increased to to 0.3881, but when we removed the 34th variable (squared) the test error increased much more significantly to 0.4987.\\
-
+We tested taking the squares of the 18th and the 34th variable individually and a curious thing happened: when we removed the 18th variable (squared) the test error increased to to 0.3881, but when we removed the 34th variable (squared) the test error increased much more significantly to 0.4987.
 \label{sec:regridgreg}
@@
 \subsection{Feature transformations}
-We tried dummy variables for all regression categorical ones.
-We could only find a significant linear correlation (using Spearman) for the 2nd class of the variable 44, 2nd and 4th class of the 47th.\\
+We tried dummy variables for all regression categorical ones. We could only find a significant linear correlation (using Spearman) for the 2nd class of the variable 44, 2nd and 4th class of the 47th.
 
@@
 
-Adding back the mentioned dummy variables 44 and 47 actually increased the test error to 0.33105. This lead the matrix $\mathbf{\tilde{X}}$ to become ill conditioned.\\
+Adding back the mentioned dummy variables 44 and 47 actually increased the test error to 0.33105. This leads the matrix $\mathbf{\tilde{X}}$ to become ill conditioned.
 
@@
 With the 18th the test error went up to 0.32137 and with the 34th it went up to 0.39349.
-So this doesn't help and we decided not tu use square root transformation.\\
+So this doesn't help and we decided not tu use square root transformation.
 
@@
 
-By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the mypoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
+By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the myPoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
 When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.3805.
@@
 Given the data description from section \ref{sec:datadescr}, this tells us that $\mathbf{\tilde{X}}$ is not ill-conditioned.\\ The range of $\mathbf{y}$ is between $-1$ and $1$.
-We also modified the values $-1$ from $\mathbf{y}$ to 0 to be consistent with the book.
+We also modified the values $-1$ from $\mathbf{y}$ to 0 to be consistent with the book of Murphy.
 We show the boxplot for $\mathbf{X}$ in Figure \ref{fig:boxplotclass}.
@@
 \label{sec:ftransclass}
-We tried to use simple logistic regression using all the variable, the estimated RMSE test error was 0.27099.
-Also we tried using sqrt on the 11th and 24th variables which actually improves our test error to 0.26779.
-We can also use squares of the variables instead of the square root and applying this alone the test error decreased to 0.26715.
+We tried to use simple logistic regression using all the variables, the estimated RMSE test error was 0.27099.
+Also we tried using square root on the 11th and 24th variables which actually improved our test error to 0.26779.
+We can also use squares of the variables instead of the square root.
+Applying this alone the test error decreased to 0.26715.
 Adding both transformations got us a decrease to a test error of 0.26474.
-We also tried to remove all categorical variables as we did in regression, but the test error increased to 0.3396, so their information was needed and cannot be ignored.\\
+We also tried to remove all categorical variables as we did in regression, but the test error increased to 0.3396, so their information was needed and cannot be ignored.
 
@@
 We could improve the test error by cross-validation.
-The Figure \ref{fig:crossval} shows the three different test error RMSE, 0-1 loss and logloss.
-We used different numbers of seed to have a better prediction for our test error.
+The Figure \ref{fig:crossval} shows the three different tests error RMSE, 0-1 loss and log loss.
+We used different numbers of seeds to have a better prediction for our test error.
 \begin{figure}
@@
 As we have seen, the cross validation helps to have a better test error.\\
-
 We obtained these RMSE for regression.
@@
 \end{description}
-We see that there is only a small improvement with the log loss error.\\
-The best error we achieved is through the logistic regression method and using the 0-1 loss error.
+We see that there is only a small improvement with the log loss error.
+The best error we achieved is through the penalized logistic regression method and using the 0-1 loss error.
 
@@
 \subsubsection*{References}
+Kevin Murphy, Machine Learning : A Probabilistic Perspective {\em The MIT Press
+Cambridge, Massachusetts
+London, England,2012}
 

### 2014-11-03 report/paper.tex
final update - fix
final update with corrections .----------------------------.

@@
 
+\subsection{Feature transformations}
+We tried dummy variables for all regression categorical ones. We could only find a significant linear correlation (using Spearman) for the 2nd class of the variable 44, 2nd and 4th class of the 47th.
+
+We tried keeping the polynomial transforms and removed all categorical variables. The test error went down to 0.33043.\\
+
+Adding back the mentioned dummy variables 44 and 47 actually increased the test error to 0.33105.
+Adding only the specifically mentioned (above) classes indicator dummy variables led the matrix $\mathbf{\tilde{X}}$ to become ill conditioned.
+
+By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the myPoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
+When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.3805.
+
+We also tried adding square root of the 18th and 34th variable, that means $\sqrt{|X_{n18}|}$ and $\sqrt{|X_{n34}|}$ of $\mathbf{X}$.
+With the 18th the test error went up to 0.32137 and with the 34th it went up to 0.39349.
+So this doesn't help and we decided not tu use square root transformation.
+
+We decided to ignore all the un-correlated variables for regression, giving a test error of 0.33173 while including the significant transformations of the 18th and 34th variables discussed before.
+We decided to opt for a simpler model because of Occam’s Razor.
+
 \subsection{Ridge Regression}
@@
 We realized that the RMSE decreased for a parameter $\lambda = 39.0694$ to a RMSE of $0.3073$. We can see that on the Figure \ref{fig:ridreggcv}.
-
-\subsection{Feature transformations}
-We tried dummy variables for all regression categorical ones. We could only find a significant linear correlation (using Spearman) for the 2nd class of the variable 44, 2nd and 4th class of the 47th.
-
-We tried keeping the polynomial transforms and removed all categorical variables. The test error went down to 0.33043.\\
-
-Adding back the mentioned dummy variables 44 and 47 actually increased the test error to 0.33105. This leads the matrix $\mathbf{\tilde{X}}$ to become ill conditioned.
-
-We also tried adding square root of the 18th and 34th variable, that means $\sqrt{|X_{n18}|}$ and $\sqrt{|X_{n34}|}$ of $\mathbf{X}$.
-With the 18th the test error went up to 0.32137 and with the 34th it went up to 0.39349.
-So this doesn't help and we decided not tu use square root transformation.
-
-We decided to ignore all the un-correlated variables for regression, giving a test error of 0.33173 while including the significant transformations of the 18th and 34th variables discussed before.
-We decided to opt for a simpler model because of Occam’s Razor.
-
-By isolating the 18th and 34th input variables, and producing polynomial transforms of them using the myPoly function, and then measuring their Pearson correlation coefficient with the output, we could see that the cube and the power 5 of the 18th variable, and powers up to 6 for the 34th variable are correlated with the output.
-When adding these feature transformations, we could lower the test error to 0.3307, which is quite lower now than the starting test error of 0.3805.
 \section{Classification}
@@
 \end{description}
-We see that there is only a small improvement with the log loss error.
+We see that there is only a small improvement with the log loss error with penalized logistic regression.
 The best error we achieved is through the penalized logistic regression method and using the 0-1 loss error.
@@
 Cambridge, Massachusetts
-London, England,2012}
-
+London, England,2012}\\
+Trevor Hastie,Robert Tibshirani,Jerome Friedman : The Elements of
+Statistical Learning. Data Mining,Inference,and Prediction. {\em Springer, 2008}\\
+Gareth James,
+Daniela Witten,
+Trevor Hastie,
+Robert Tibshirani:An Introduction to Statistical Learning.{\em Springer, 2013}
 \end{document}

### 2014-11-03 report/paper.tex
report final

@@
 We used different numbers of seeds to have a better prediction for our test error.
+For RMSE and 0-1 loss, we tried to change the range of the $\lambda$ parameter but we got the same curve.
 \begin{figure}
@@
 \end{description}
-For the regression, we saw that the best method was Ridge Regression. The best parameter was $\lambda =  39.0694$ and we achieved a test error of 0.3073.
+For the regression, we saw that the best method was Ridge Regression.
+The best parameter was $\lambda =  39.0694$ and we achieved a test error of 0.3073.
+This is obtained by removing the categorical, removing the un-correlated variables and using the cube of the 18th and the degree 6 with the 34th variable.
 
@@
 \end{description}
+This is obtained by using the squares and the square root of the 11th and 24th variables.
 With logistic regression we obtained different results depending the type of error we computed. \begin{description}

### 2014-11-20 Project 1/report/paper.tex
PROJECT II STARTING

--- 2014-11-03 report/paper.tex
@@
 The best parameter was $\lambda =  39.0694$ and we achieved a test error of 0.3073.
-This is obtained by removing the categorical, removing the un-correlated variables and using the cube of the 18th and the degree 6 with the 34th variable.
+This is obtained by removing the categorical, removing the un-correlated variables and using the cube of the 18th and up to the degree 6 with the 34th variables.
 

### 2014-11-20 Project 1/script.m

@@ -1,0 +1,182 @@
+% Comments and observations made by youssef
+
+clc
+clear all
+close all
+
+%load('MexicoCity_regression.mat')
+load('MexicoCity_classification.mat')
+
+N = length(y_train); % data size
+D = size(X_train, 2); % dimensionality
+
+%% plotting the histograms of the variables
+noOfVertPlots = floor(sqrt(D));
+noOfHorFigures = ceil(D/noOfVertPlots);
+
+figure;
+suptitle('histogram of all the input variables')
+for i = 1:D
+    subplot(noOfVertPlots, noOfHorFigures, i)
+    hist(X_train(:, i))
+    title(num2str(i))
+end
+
+figure;
+hist(y_train);
+title('hisogram of output variable')
+
+%% categorical variable detection
+
+noOfVariableValues = zeros(D, 1);
+
+% SUPPOSING categorial variables have AT MOST 15 classes
+maxClasses = 15;
+
+for i=1:D
+    test = unique(X_train(:, i));
+    noOfVariableValues(i) = length(test);
+end
+
+catVars = find(noOfVariableValues < maxClasses);
+
+%% normalizing the features
+
+% for regression variables
+% normalize features (store the mean and variance)
+meanX = mean(X_train);
+
+% 42 first features seem to be gaussian distributed, can be normalized
+normInputVars = setdiff(1:D, catVars);
+
+% normalizing the input variables
+X_train(:, normInputVars) = X_train(:, normInputVars) - repmat(meanX(normInputVars), N, 1);
+stdX = std(X_train);
+X_train(:, normInputVars) = X_train(:, normInputVars)./repmat(stdX(:, normInputVars), N, 1);
+
+% normalizing the output variable, comment out for classification
+y_train = (y_train-mean(y_train))/std(y_train);
+
+% Form (y,tX) to get regression data in matrix form
+tX = [ones(N,1) X_train(:, :)];
+
+%% correlation & normal distribution analysis
+
+% taking the 18th and 34th variables, and any polynomial transform of them,
+% and checking their pearson correlation with the output
+
+txSpecial = [myPoly(X_train(:, 18),5) myPoly(X_train(:, 34),10) ];
[…]

### 2014-11-27 Project 2/hw5/hw5.tex
hw5 extracted

\documentclass[11pt,a4paper,notitlepage]{article}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{enumerate}
\usepackage[left=2.00cm, right=2.00cm, top=2.00cm, bottom=2.00cm]{geometry}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{listings}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage{arydshln}
\lhead{Big Data}

\chead{Homework 5}
\rhead{Kevin Serrano, 204141}
\pagestyle{fancy}
\author{Kevin Serrano}
\title{Homework 5}

\begin{document}
\maketitle
\section*{Part 1}
If we multiply $R$ by $R^t$, we obtain a similarity matrix. Row $i$ shows the number of items that are liked by user $i$ and $j$, where the user $j$ is the $j$-th column.\\
\underline{Example:}\\
Let $m = 3$ and $n = 5$. Define $$R = \left[\begin{matrix}
0 & 1 &    0  &   1 &    1 \\
     1  &   0   &  0 &     1  &   0\\
     0   &  1   &  1  &   1   &  0
\end{matrix}\right]$$ Then $$R \cdot R^t = \left[ \begin{matrix}
     3  &   1   &  2\\
     1  &   2  &   1\\
     2  &   1 &    3
\end{matrix}\right]$$
That means user $1$ likes $3$ items in common with himself (obviously), 1 with user 2 and 2 with user 3. \\

By the definition of the cosine similarity, we need to divide each $(i,j)$ by the norm of the number of item liked by the user $i$ and $j$ (the norm is the square root here). To applied this to the all matrix, we just need to divide $R\cdot R^t$ by the sqrt of $P$ : $\sqrt{P}$.\\
Thus $$S_U = \sqrt{P}^{-1} \cdot R \cdot R^t \cdot \sqrt{P}^{-1}$$
For the example we obtain $$\frac{R \cdot R^t}{\sqrt{P} \cdot \sqrt{P}} =\sqrt{P}^{-1} \cdot R\cdot R^t\cdot  \sqrt{P}^{-1}= \left[\begin{matrix}
    1.0000  &  0.4082 &   0.6667\\
    0.4082  &  1.0000 &   0.4082\\
    0.6667  &  0.4082 &   1.0000
\end{matrix}\right]$$
where for example $0.4082 = \frac{1}{\sqrt{3} \cdot \sqrt{2}}$\\

By the same way, we can compute the cosine similarity for items as follow :\\
$R^t\cdot R$ gives us the number of user who like item $i$ and $j$. With the example it is $$R^t\cdot R =\left[ \begin{matrix}
     1  &   0   &  0&     1    & 0\\
     0  &  2    & 1  &   2    & 1\\
     0  &   1   &  1  &   1  &   0\\
     1  &  2   &  1    & 3  &   1\\
     0  &   1 &    0    & 1&     1
\end{matrix}\right]$$
That means for example that 2 users like item 2 and 4.
Now we can just divide by the sqrt of $Q$ to get the cosine similarity between items $$S_I = \sqrt{Q}^{-1} \cdot R^t \cdot R \cdot \sqrt{Q}^{-1}$$
For the example we get $$\sqrt{Q}^{-1} \cdot R^t\cdot R \cdot \sqrt{Q}^{-1} = \left[\begin{matrix}

    1.0000  &       0    &     0&    0.5774    &     0\\
         0   & 1.0000   & 0.7071 &   0.8165   & 0.7071\\
         0    &0.7071  &  1.0000  &  0.5774  &       0\\
    0.5774    &0.8165 &   0.5774   & 1.0000 &   0.5774\\
         0    &0.7071&         0    &0.5774&    1.0000
\end{matrix}\right]$$
\section*{Part 2}
\begin{enumerate}[(a)]
\item  \textbf{User-user collaborative filtering}\\

We have $$X = \left[\begin{matrix}
x_{11} & x_{12} & \cdots & x_{1n}\\
x_{21} & x_{22} & \cdots & x_{2n}\\
\vdots & \vdots & \ddots & \vdots\\
x_{m1} & x_{m2} & \cdots & x_{mn}
\end{matrix}\right]$$ where $$x_{ij} = \sum_{i=1}^{m} r_{ij} \cdot su_{ij}$$ and $r_{ij} \in R, su_{ij} \in S_U$. Need to be carefull for matrix dimension, thus we get $$X = S_U \cdot R = (\sqrt{P}^{-1} \cdot R \cdot R^t \cdot \sqrt{P}^{-1}) \cdot R$$

\item \textbf{Item-item collaborative filtering}\\
For the same reason, we have $$x_{ij} = \sum_{j=1}^{n} r_{ij} \cdot si_{ij}$$ with $r_{ij} \in R, si_{ij} \in S_I$. Thus we get $$X = R \cdot S_I = R \cdot (\sqrt{Q}^{-1} \cdot R^t \cdot R \cdot \sqrt{Q}^{-1})$$
\end{enumerate}

\section*{Part 3}
Let's apply the algorithms for user Bob ($200$-th row). I used Matlab for the implementation. See the code for more details.
\begin{enumerate}[(a)]
\item With the user-user collaborative filtering, I obtain these shows with these scores\\

\begin{tabular}{c c}
TV-show & score\\
\hline
    'FOX 28 News at 10pm' & 882.8699\\ 
    'Family Guy' & 828.5152\\
    'NBC 4 at Eleven' & 780.2172\\
    '2009 NCAA Basketball Tournament'& 765.7426\\
    'Access Hollywood'& 750.3544\\
\end{tabular}

\item With the item-item collaborative filtering, I obtain these shows with these scores\\

\begin{tabular}{c c}
TV-show & score\\
\hline
    'FOX 28 News at 10pm' & 23.8006\\    
    'NBC 4 at Eleven' & 22.6289\\
    'Access Hollywood'& 22.6210 \\
    'Family Guy'& 22.6071    \\
    'Two and a Half Men'& 22.0486
\end{tabular}

\item It is easy to check the precision. For both filtering, I get a precision of 1 for $k=5$. That means Bob have watched all movies that are proposed by the user-user or item-item filtering.
\end{enumerate}
\end{document}

### 2014-11-27 Project 2/hw5/hw5.m

@@ -1,0 +1,92 @@
+%% Load file
+users_shows = fopen('user-shows.txt','r');
+sizeR= [563 9984];
+formatSpec = '%f';
+R = fscanf(users_shows,formatSpec,sizeR)';
+TS = textread('shows.txt','%q');
+%% 100 TV shows and Bob
+S= TS(1:100);
+
+bob = R(200,:);
+%% Test case (from my pdf, used for debug my code)
+testR = [0 1 0 1 1 ; 1 0 0 1 0; 0 1 1 1 0];
+testP = [3 0 0 ; 0 2 0 ; 0 0 3];
+testQ = [1 0 0 0 0; 0 2 0 0 0 ; 0 0 1 0 0; 0 0 0 3 0;0 0 0 0 1];
+
+%% Compute P and Q
+[m,n] = size(R);
+for i = 1:m
+    row = R(i,:);
+   diagP(i) = sum(row);
+end 
+P = diag(diagP);
+for(i = 1:n)
+    col = R(:,i);
+    diagQ(i) = sum(col);
+end
+Q = diag(diagQ);
+
+%% check if diagQ and diagP has no zero
+any(0==diagP)
+any(0==diagQ)
+%% Compute inverse of sqrt P and Q
+% Because P,Q are diagonal with non-zero element in the diagonal, the inverse is simply the inverse of the
+% diagonal
+for i = 1:m
+    invSqrtDiagP(i) = sqrt(1/diagP(i));
+end
+invSqrtP = diag(invSqrtDiagP);
+
+%%
+for i = 1:n
+    invSqrtDiagQ(i) = sqrt(1/diagQ(i));
+end
+invSqrtQ = diag(invSqrtDiagQ);
+%%
+
+RRt = R*R';
+RtR = R'*R;
+%% Compute Su and Si as in the pdf: This take a few minutes
+Sup1 = bsxfun(@times,diag(invSqrtP),RRt); % optimization for diagonal matrix multiplication (thanks stack overflow)
+Su = Sup1*invSqrtP;
+%Si = invSqrtQ*RtR*invSqrtQ;
+Sip1 = bsxfun(@times,diag(invSqrtQ),RtR);
+Si = Sip1*invSqrtQ;
+
+%% user-user clollaborative (part 3.a)
+X = Su * R;
+% get the top-k values and their indices
+k = 5;
+[sortX,sortIdx] = sort(X(200,1:100),'descend'); % sort the first 100 shows scores for user bob
+maxKVal = sortX(1:k)
[…]

### 2014-11-27 Project 2/kevin/mr_play.m
play kevin

@@ -1,0 +1,88 @@
+%% Load file
+close all; clear all;
+load('../songTrain.mat');
+ua = Ytrain; % users-artists
+
+%% Normalize per user (rows)
+rowMax = max(ua,[],2);
+rowMax(rowMax==0) = 1;
+%%
+rowrep = repmat(rowMax,1,size(ua,2));
+ua = ua./rowrep;
+%% Compute P and Q
+[m,n] = size(ua);
+for i = 1:m
+    row = ua(i,:);
+   diagP(i) = sum(row);
+end 
+P = diag(diagP);
+for(i = 1:n)
+    col = ua(:,i);
+    diagQ(i) = sum(col);
+end
+Q = diag(diagQ);
+
+%% check if diagQ and diagP has no zero
+any(0==diagP)
+any(0==diagQ)
+%% Compute inverse of sqrt P and Q
+% Because P,Q are diagonal with non-zero element in the diagonal, the inverse is simply the inverse of the
+% diagonal
+for i = 1:m
+    invSqrtDiagP(i) = sqrt(1/diagP(i));
+end
+invSqrtP = diag(invSqrtDiagP);
+
+%%
+for i = 1:n
+    invSqrtDiagQ(i) = sqrt(1/diagQ(i));
+end
+invSqrtQ = diag(invSqrtDiagQ);
+%%
+
+RRt = ua*ua';
+RtR = ua'*ua;
+%% Compute Su and Si as in the pdf: This take a few minutes
+Sup1 = bsxfun(@times,diag(invSqrtP),RRt); % optimization for diagonal matrix multiplication (thanks stack overflow)
+Su = Sup1*invSqrtP;
+%Si = invSqrtQ*RtR*invSqrtQ;
+%Sip1 = bsxfun(@times,diag(invSqrtQ),RtR);
+%Si = Sip1*invSqrtQ;
+
+%% user-user clollaborative (part 3.a)
+X = Su * ua;
+% get the top-k values and their indices
+k = 5;
+[sortX,sortIdx] = sort(X(200,1:100),'descend'); % sort the first 100 shows scores for user bob
+maxKVal = sortX(1:k)
+maxKindx = sortIdx(1:k);
+topKshows = S(maxKindx) % get the name of the shows
+%% movie-movie collaborative (part 3.b)
+Y = ua * Si;
+k = 5;
+% get the top-k values and their indices
+[sortY,sortIdy] = sort(Y(200,1:100),'descend');
[…]

### 2014-11-27 Project 2/kevin/mr_play.m
mr_play kevin update 2

@@ -1,7 +1,7 @@
 %% Load file
 close all; clear all;
 load('../songTrain.mat');
-ua = Ytrain; % users-artists
+ua = Ytrain; % users-artists 
 
 %% Normalize per user (rows)
 rowMax = max(ua,[],2);

### 2014-12-10 Project 2/Youssef/youssefplaying2.m
modified kevin's functions

@@ -66,7 +66,7 @@ LRTPR = TPRs(Te.y, LRPred);
 
 %% Fitting using penalized logistic regression - simple run
 
-lambda = 1/50;
+lambda = 44;
 betaPLR = penLogisticRegression(Tr.y,Tr.tX,alpha,lambda);
 temp2 = Te.tX*betaPLR;
 PLRPred = probEstimate(temp2)*2-1;

### 2014-12-10 Project 2/predictSVMoutput.m

@@ -1,12 +1,11 @@
-function [ binaryScore ] = predictSVMoutput(Xte, SVMModelTr )
+function [ scores ] = predictSVMoutput(Xte, SVMModelTr )
 
 %% predict new data
 fprintf('Predict new data...\n');
 [~, scores] = predict(SVMModelTr,Xte);
 
 %%
-binaryScore = scores(:,1) - scores(:,2);
-
+%finalScore = scores(:,1) + scores(:,2);
 
 end
 

### 2014-12-10 Project 2/trainSVM.m

@@ -1,10 +1,16 @@
 function [ SVMModelTr ] = trainSVM( YTr, XTr, kernel, C )
-%WRAPPERSVM Summary of this function goes here
-%   Detailed explanation goes here
 
-%% create the model
+% create the model
 fprintf('Creating the model..\n');
-SVMModelTr = fitcsvm(XTr,YTr, 'boxconstraint', C, 'kernel_function', kernel);
 
+if isempty(C)
+    C = 1;
 end
 
+if isempty(kernel)
+    kernel = 'linear';
+end
+
+SVMModelTr = fitcsvm(XTr,YTr, 'BoxConstraint', C, 'KernelFunction', kernel);
+
+end

### 2014-12-11 Project 2/report/paper.tex
report update

\documentclass{article} % For LaTeX2e
% We will use NIPS submission format
\usepackage{nips13submit_e,times}
% for hyperlinks
\usepackage{hyperref}
\usepackage{url}
% For figures
\usepackage{graphicx} 
\usepackage{subfigure} 
% math packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsopn}
\usepackage{ifthen}
\usepackage{natbib}

\usepackage{caption}

\title{Project-II by Group MexicoCity}

\author{
Kevin Serrano\\EPFL\\
\texttt{kevin.serrano@epfl.ch} \And Youssef El Baba\\EPFL\\
\texttt{youssef.baba@epfl.ch} \And Alexandre Helfre\\EPFL\\
\texttt{alexandre.helfre@epfl.ch}
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\nipsfinalcopy

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

\section{Exploratory Data analysis}
\label{sec:datadescr}
For the person detection dataset, there are 8545 images and a feature vector of dimensionality 9360 for each. The images are 3-channel (RGB) images, and the features are simply a vectorized form the image HOG features. Clearly this is a fat data matrix, even more so if we are going to split it into train/test datasets. Anyway, considering the huge number D, we are tempted to do some sort dimensionality reduction if we are to avoid extremely complex models and overfitting. This hints to the possibility of using PCA for this dataset. Or alternatively we can use neural networks, with the help of the Deep Learning Toolbox (kindly indicated by the course staff). We can also try using simple tools like $k-$means with $K = 2$ or SVM.

On initial visual inspection of the images and their corresponding features, it does seem that the HOG features for images with people are of a somewhat different nature. The various blocks containing the $(W \cdot W)$ feature blocks seem to have a much more varied orientation than for the images without people.

As to the music recommendation dataset, there are a total of 1774 users and 15082 artists. Each pair of (user, artist) is a data element, and the corresponding listening count number is the data.. If we change our convention to pairs of (user, user) we get another feature vector of size D=1 that is the friendship relation (currently given as a binary number, indicating a friendship existance). The format of the data set is peculiar with respect to usual matrices containing N rows of D columns (features) each. We can see that some users listen practically solely for a single artist with a high count when inspecting the provided $Y_{\text{train}}$ matrix.

On a different note, it seems daunting at first to predict specific count numbers for each (user,artist) pair, as usually the recommendation in such systems are the k most “likeable” artists for a specific user. One way to go is to predict, using a computed probability for each (user, artist), a sort of score that is, a count that will be simply the maximum number of counts of the user when the score is 1, and adjusted linearly with the score as the coefficient.

\section{People Detection}
\subsection{Neural Networks}
When playing with the person detection code given to us, using the deep learning toolbox, the training error obtained was 0.745, which is approx. 75%, not bad at all for a first trial. Since the number of data points is small, a non-zero value of the dropout rate might help. We should also consider L2 penalization (similar to lambda used in ridge regression) in the late stages of the project to reduce the effects of overfitting. We will then try to measure the average test TPR with k-fold cross-validation.

Continuing with neural networks testing, we tried adding a dropout fraction of 0.3 and see if it could improve the error rates. The stable (after 25 epochs) mean-square error (obtained using the native plot of the Deep Learning Toolbox nntrain function) for the methods without and then with dropout was 0.005774 and 0.006937 respectively. Oddly enough though, average TPRs for these methods are 0.75 and 0.787 respectively. A second run with another seed for the Matlab random number generator gave other final MSEs and TPRs, this is because NN are very sensitive to the initial seed. So the seed for all further trials was manually set to 8339 (given by the course staff) . Adding a dropout seems to enhance our TPR,, but gains are somewhat mitigated for bigger FPRs. The batchsize for these methods (a parameter in the Deep Learning toolbox) was also tampered with, when we tried to take the whole training/data as a batch (batch size ~4272), however this turned out to be a bad idea as the convergence of the MSE cost (on the toolbox’s training graphs) only attained 0.014 at the lowest level, much higher than with a batchsize of 100. Then we tried changing the internal activation functions from tanh to sigmoids, with an improvement of the MSE / TPR to 0.002975 and 0.795, which was quite remarkable. The last 2 variants to be tried were simple NN with $L_2$ weight penalty equal to 1/1000 and simple NN with both sigmoid activations and a dropout of 0.3, which respectively gave MSE / TPR values 0.006758 / 0.772 of 0.00404 / 0.844. These results

In addition to that, logistic regression was given a try, and it gave a TPR of 0.7521. Then Penalized logistic regression with a lambda of 1/50 also gave the same TPR. The $\lambda$ for the logistic regression clearly wasn’t adequate.

One clear conclusion that can be drawn here is that internal sigmoid activation functions, instead of tanh, help a lot. They are going to be used for the final model then (if using NN).

Clearly, each of these additions to the basic NN model improves things, on its own, a bit. However when they are applied together, things improve considerably (as indicated by the last cited and quite high TPRs). Of course, the values for these parameters (dropout and $L_2$ penalty) were chosen arbitrarily. Similarly, Penalized Logistic regression can have a different optimal lambda, which should be cross-validated, and might result in higher TPR.

\subsection{Penalize Logistic regression (cross-validation for $\lambda$}
Penalized logistic regression was cross-validated (for the lambda parameter) to see which setting gives the best estimated TPR. The error metric taken was 1-TPR, which made simple sense. $\lambda$ values between $10^{-3}$ and $10^2$ were tested, 15 of them, and using 15 different seeds. So a total of $15 \cdot 15=225$ gradient descent runs for PLR were tried, giving the following plot (Figure \ref{fig:TEPLR})
It seems that, roughly, the more we increase lambda, the better our test error equivalent (1-TPR, which is actually the False Negative Rate) is.
\begin{figure}[h!]
\centering
\includegraphics[scale=0.6]{images/PLRKCVplot.png}
\caption{Test error PLR with various $\lambda$}
\label{fig:TEPLR}
\end{figure} For $\lambda$ = 44, we deem it the test error no longer decreases significantly and it’s sufficiently good at 0.1849 (this corresponds to a TPR of 0.8151). This would be the best possible variant in the Logistic regression family to use. It remains to be seen at this point, however, if other ML methods can do better, especially after all their variants have been cross-validated (NN with various dropouts and $L_2$ weight penalties for instance). (red curves show the 1-TPR for test set, using model trained on a different training set, for each lambda, each curve being the result of a different seed, the bold curve being the average).

\subsection{$K-$means}
We then tried using $k-$means on the same train and test sets used for the previous methods, with identical normalization (essentially how the provided demo script did the split). The MATLAB $K-$means is particular in the sense that it does not take the actual (true) labels (person / no person) of each point (image) in account when computing its assignment, it simply renames the classes as $1, 2, 3, \cdots, K$. What it does is it simply clusters in a best effort manner the points, and assigns random labels to them. We tried at first to simply compute the MATLAB labels on the test data and reassign them to $-1$ and 1, which gave a TPR of 0.0057 (or 0.57 \%), very low for any practical purpose to say the least. We then tried to re-wrap the $k-$means function in another one generating a "model" (centroids + labels of the centroids computed as the mean of the labels of their assigned points) which now uses the true labels of the points. And another function which, as is typical of other methods, predicts the assignments of new points (using euclidean distance to centroids). This now resembled classical model based methods, however it also failed miserably with a TPR of $5.8*10^{-4}$. This can be explained by the fact that the cluster of no-person images is much, much larger than the one with person-containing images, hence when assigning the labels to the cluster means they consistently get the label "no-person" making the result a static label $-1$. As taking the mean label of the cluster points and assigning it to the cluster mean is an integral part of the $K-$means algorithm, this ML method is probably not the one to use here.

\subsection{Support Vector Machine}
The next method we tried to apply on the people detection dataset was SVM. We used the \texttt{fitcsvm} and \texttt{predict} functions of MATLAB, to be able to obtain posteriors probabilities and not just hard assignments. We tested plain old SVM, with a linear kernel, and by taking a slight transform of the posterior we were able to obtain a TPR of 80.59 \%. We tried varying the $C$ constant parameter, trying values of 0.1, 0.5, 2 and 5, however they all gave the same TPR. We then tried using an RBF kernel, with $\sigma$ values of 0.5, 1 and 2, but they all gave a very low, consistent TPR of 0.0582 \%. This is very similar to what we get by simply predicting a $-1$ (no-person) for all the images. The prospects for the basic variant of this method provided good performance, but its lack of flexibility for further improvement makes us believe we should opt for another ML method for this dataset instead.

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{images/TPRplot.png}
\caption{Comparison of different methods for People detection}
\label{fig:pdplot}
\end{figure}
\section{Music recommendation}
\subsection{Histogram}
Following is a the histogram of the counts for the Music recommendation dataset (Figure \ref{fig:histC}). The max count is 352698, which is huge. Moreover the number of counts $>$ 5000 is rather small. That’s why we’re only showing the bins for the counts that are smaller or equal to 5000, the rest being generally present once or twice at most.

After observing the histogram of the counts, as proposed by the course staff, we can see that despite the high maximal count, most of the data is concentrated in the low-count range (1-3500). Going out on a limb here we might consider any count quite outside of this interval to be an outlier, however things are a bit nuanced: if we start removing outliers from the data (which is already sparse) we run into the problem of deciding how far the point has to be from this interval to consider it as such. That decision is very difficult to make, especially considering that music listening - as any human-  behaviour cannot be expected to always follow mathematical rules. \textcolor{red}{What about poisson distribution for count?}

A second conjecture is that, inside this interval, most counts are still concentrated at the lowest range. It’s as if the histogram of the counts decays exponentially fast, which is why taking the logarithm (of the counts) would change the $x$ axis scaling in some sorts and produce a more bell shaped curve, i.e. a Gaussian. This is experimentally confirmed by taking the histogram of the counts (full counts, not restricted to the interval, see Figure \ref{fig:histC}) and then taking the histogram of the log of the counts, which does correspond to a Gaussian, as shown in Figure \ref{fig:loghist}. This shows potential for such a transform to be used before applying ML methods, especially those one imposing Gaussian-distributed restrictions on their data.
\begin{figure}
\centering     %%% not \center
\subfigure[Histogram of counts]{\label{fig:histC}\includegraphics[width=0.49\textwidth]{images/countHist.png}}
\subfigure[Histogram of log transformed counts]{\label{fig:loghist}\includegraphics[width=0.49\textwidth]{images/MRhistlog.png}}
\caption{Histograms}
\end{figure}

\subsection{PCA/SVD}
We ran the PCA algorithm on the full-format $Y_{train}$ data. Nevertheless the ALS algo (PCA with the ALS option) does not work with this matrix, as it seems that $Y_{train}$ is a too large a matrix even if it is sparse in practice.

Therefore we applied the SVD decomposition on $Y_{train}$ to see if we can reduce in practice reduce the dimensionality of our problem. To illustrate that possibility, following is the plot (Figure \ref{fig:modk}) of the cumulative fraction of variance explained vs the modes. (For mode 1 we use singular value $s_1$ and its corresponding vector, for mode 2 we use $s_1$ and $s_2$ and their etc, increasing the rank of our approximation). The result can be summarized in that the 900 most significant vectors (highest singular value) explains 95\% of the data’s variance, which means they can be roughly sufficient to use as a basis for training ML models and predictions.
\begin{figure}
\centering
\includegraphics[width=1\textwidth]{images/svdvariancemodes.png}
\caption{Cumulative variance fraction explained}
\label{fig:modk}
\end{figure}

\subsection{Basic mean-based predictions}
One of the simplest prediction method is to use averages to predict. The first thing we have done is to compute the average listening per artist on our training set.  Then using only the positive values in our test set, we can compute the RMSE between our prediction and the real values. If we don’t apply any transformation to the data, we obtain a RMSE of 6400. We can see that this is quite a high number but nevertheless expected since the RMSE supposes an Gaussian input and that the average method is not a good one.

Then we applied the log transformation to obtain a more Gaussian input as seen in Figure \ref{fig:loghist}. We then compute the RMSE based on the same principle as before and rescale the result with the exponential to obtain an RMSE of 4130, which is already much better.

\subsection{Collaborative-filtering}

\section{Summary}

\subsubsection*{Acknowledgments}

\subsubsection*{References}
Kevin Murphy, Machine Learning : A Probabilistic Perspective {\em The MIT Press
Cambridge, Massachusetts
London, England,2012}\\
Trevor Hastie,Robert Tibshirani,Jerome Friedman : The Elements of
Statistical Learning. Data Mining,Inference,and Prediction. {\em Springer, 2008}\\
Gareth James,
Daniela Witten,
Trevor Hastie,
Robert Tibshirani:An Introduction to Statistical Learning.{\em Springer, 2013}
\end{document}

### 2014-12-11 Project 2/analyzeKCVresults.m
update to SVM prediction

@@ -8,11 +8,11 @@ avgErrorTe = mean(errorTe);
 avgErrorTr = mean(errorTr);
 
 figure;
-plot(parameterValues, errorTe,'r-','color',[1 0.7 0.7]);
+plot(parameterValues', errorTe','r-','color',[1 0.7 0.7]);
 hold on
-plot(parameterValues, errorTr,'b-','color',[0.7 0.7 1]);
-plot(parameterValues, avgErrorTe ,'r-','linewidth', 3); hold on
-plot(parameterValues, avgErrorTr,'b-','linewidth', 3); xlabel(parameterName);
+plot(parameterValues, errorTr','b-','color',[0.7 0.7 1]);
+plot(parameterValues, avgErrorTe' ,'r-','linewidth', 3); hold on
+plot(parameterValues, avgErrorTr','b-','linewidth', 3); xlabel(parameterName);
 ylabel(errorName)
 title(givenTitle);
 

### 2014-12-11 Project 2/genericKCV.m

@@ -53,19 +53,15 @@ for s = 1:noOfSeeds
             % this ASSUMES that all input variables are gaussian
             % distributed
             switch fittingFunctionType
-                case 9
+                case 10
                     % these cases correspond to the cases where inputX = tX
                     % ( has a column of ones in the beginning
-                    
-                    % start counting at second column the input variables, because first one is a dummy column of 1's
-                    nonNormalInputVars = nonNormalInputVars + 1;
                     
                     [tXTr, mu, sigma] = zscore(tXTr); % train, get mu and std
                     tXTe = normalize(tXTe, mu, sigma);  % normalize test data
                    
                 otherwise
-                    % these cases correspond to the cases where inputX = tX
-                    % ( has a column of ones in the beginning
+                    % these cases correspond to the cases where inputX = X
                     
                     [normXTr, mu, sigma] = zscore(tXTr(:, 2:end)); % train, get mu and std
                     normXTe = normalize(tXTe(:, 2:end), mu, sigma);  % normalize test data
@@ -86,12 +82,14 @@ for s = 1:noOfSeeds
                     end
                 case 4 % only penalized log. reg. used 4 arguments
                     beta = fittingFunction(yTr, tXTr, alpha, parameter(i));
-                case 9 % only NN takes so many input arguments !
+                case 10 % only NN takes so many input arguments !
                     
                     dimensions = constantNNoptions.dimensions;
                     noEpochs = constantNNoptions.noEpochs;
                     batchSize = constantNNoptions.batchSize;
                     learningRate = constantNNoptions.learningRate;
+                    
+                    activFun = constantNNoptions.activationFun;
                     
                     % the data size needs to be a multiple of the batchsize
                     numTrSampToUse = batchSize * floor( size(tXTr) / batchSize);
@@ -100,14 +98,16 @@ for s = 1:noOfSeeds
                     tXTr = tXTr(1:numTrSampToUse,:);
                     yTr = yTr(1:numTrSampToUse);
                     
+                    dimensions = [size(tXTr, 2) dimensions];
+                    
                     if isfield(constantNNoptions, 'dropout')
                         % we're varying the L2 weight and keeping the
                         % dropout constant in this case
                         
                         dropout = constantNNoptions.dropout;
                         
-                        NN = fittingFunction( tXtr, yTr, dimensions, noEpochs,...
-                            batchSize, plotFlag, learningRate, dropout, parameter(i) );
[…]

### 2014-12-11 Project 2/predictSVMoutput.m

@@ -1,11 +1,15 @@
-function [ scores ] = predictSVMoutput(Xte, SVMModelTr )
+function [ finalScore ] = predictSVMoutput(Xte, SVMModelTr )
 
-%% predict new data
+% predict new data
 fprintf('Predict new data...\n');
 [~, scores] = predict(SVMModelTr,Xte);
 
-%%
-%finalScore = scores(:,1) + scores(:,2);
+% Assuming -1 and +1 labels, as is the case for our project
+% finalScore = scores(:, 2)-min(scores(:, 2));
+% finalScore = finalScore./max(finalScore);
+% finalScore = 2*finalScore-1;
+
+finalScore = scores(:, 2)
 
 end
 

### 2014-12-11 Project 2/trainSVM.m

@@ -1,4 +1,5 @@
-function [ SVMModelTr ] = trainSVM( YTr, XTr, kernel, C )
+function [ SVMModelTr ] = trainSVM( YTr, XTr, kernel, kernelScale, C )
+% use caution with RBF: it completely breaks things !
 
 % create the model
 fprintf('Creating the model..\n');
@@ -11,6 +12,10 @@ if isempty(kernel)
     kernel = 'linear';
 end
 
-SVMModelTr = fitcsvm(XTr,YTr, 'BoxConstraint', C, 'KernelFunction', kernel);
+if isempty(kernelScale)
+    kernelScale = 1;
+end
+
+SVMModelTr = fitcsvm(XTr,YTr, 'BoxConstraint', C, 'KernelFunction', kernel, 'KernelScale', kernelScale);
 
 end

### 2014-12-11 Project 2/predictSVMoutput.m
update to predictSVMoutput

@@ -9,7 +9,7 @@ fprintf('Predict new data...\n');
 % finalScore = finalScore./max(finalScore);
 % finalScore = 2*finalScore-1;
 
-finalScore = scores(:, 2)
+finalScore = scores(:, 2);
 
 end
 

### 2014-12-11 Project 2/analyzeKCVresults.m
some graph + code updates

@@ -1,5 +1,5 @@
 function [bestParameter, achievedErrorTe ] = analyzeKCVresults( errorTe,...
-    errorTr, parameterName, parameterValues, givenTitle, errorName )
+    errorTr, parameterName, parameterValues, givenTitle, errorName, modeFunction )
 % plots the results from cross-validating the given parameter, for
 % inspecttion and also retrieves the optimal parameter value and returns it
 
@@ -18,7 +18,8 @@ title(givenTitle);
 
 % computes the best parameter (usually a penalizazion term lambda) that
 % achieves the best test Error, returns both
-[achievedErrorTe, bestParameter] = min(avgErrorTe);
+
+[achievedErrorTe, bestParameter] = modeFunction(avgErrorTe);
 bestParameter = parameterValues(bestParameter);
 
 end

### 2014-12-11 Project 2/Youssef/svmKCV.m
added svm KCV script

@@ -1,0 +1,53 @@
+% script for cross-validating SVM
+
+% preparing data
+clear all;
+close all;
+clc
+
+projectDir = '/users/youssef/Documents/Matlab/PCML/Projects/Project 2';
+addpath(genpath(projectDir));
+rmpath([projectDir '/Piotr toolbox'])
+cd(projectDir);
+
+% Load both features and training images
+load train_feats;
+%load train_imgs;
+N = length(labels);
+
+% Generate feature vectors (so each one is a row of X)
+disp('Generating feature vectors..');
+D = numel(feats{1});  % feature dimensionality
+X = zeros([N D]);
+
+for i=1:N
+    X(i,:) = feats{i}(:);  % convert to a vector of D dimensions
+end
+
+y = labels;
+
+normalizeFlag = 1;
+K = 4;
+noOfSeeds = 3;
+constantSVMoptions.kernel = 'rbf';
+
+%% KCVating the kernel scale (sigma for rbf)
+
+%choose the no of sigmas carefully, with the noOfSeeds & K in mind, because
+% the code is going to train once and predict twice (test, train)
+% noOfSeeds*K*noOfSigmas models. For rbf kernels this is going to take
+% WHILE (test to see how much 1 model takes to train, if it's t minutes than the
+% code is going to take t*noOfSeeds*K*noOfSigmas/60 hours !! For 10 minutes
+% with the current parameters if it's 10 minutes per model it's 10*3*4*4/60
+% = 8 hours. Measure time and change parameters yourself
+
+noOfSigmas = 3;
+sigmas = logspace(-1, 3, noOfSigmas);
+constantSVMoptions.C = []; % indicates that this isn't the parameter to genericKCV
+
+[errorsTr, errorsTe] = genericKCV( y, X,...
+   @trainSVM, @predictSVMoutput, @TPRs, sigmas, K, [], noOfSeeds,...
+   [], constantSVMoptions, normalizeFlag);
+
+% it saves the results directly, these are to be shared
+save('SVM KCV results (light)','errorsTr', 'errorsTe', 'noOfSeeds', 'K', 'sigmas', 'constantSVMoptions');

### 2014-12-11 Project 2/Youssef/youssefplaying2.m

@@ -1,8 +1,8 @@
 % youssef's playing with the data script, project 2
 
-clear all;
-close all;
-clc
+% clear all;
+% close all;
+% clc
 
 projectDir = '/users/youssef/Documents/Matlab/PCML/Projects/Project 2';
 addpath(genpath(projectDir));

### 2014-12-11 Project 2/genericKCV.m

@@ -1,6 +1,6 @@
 function [errorsTr, errorsTe] = genericKCV( y, inputX,...
-    fittingFunction, errorFunction, parameter, K, alpha, noOfSeeds,...
-    constantNNoptions, normalizeFlag)
+    fittingFunction, predictionFunction, errorFunction, parameter, K, alpha, noOfSeeds,...
+    constantNNoptions, constantSVMoptions, normalizeFlag)
 % generic cross-validation function used to estimate the best parameter
 % (the one giving the best Test/Train error pair (more emphasis on the
 % latter). It can be used with any fitting method (e.g. least squares) and
@@ -53,7 +53,7 @@ for s = 1:noOfSeeds
             % this ASSUMES that all input variables are gaussian
             % distributed
             switch fittingFunctionType
-                case 10
+                case {5, 10}
                     % these cases correspond to the cases where inputX = tX
                     % ( has a column of ones in the beginning
                     
@@ -82,6 +82,23 @@ for s = 1:noOfSeeds
                     end
                 case 4 % only penalized log. reg. used 4 arguments
                     beta = fittingFunction(yTr, tXTr, alpha, parameter(i));
+                case 5 % only SVM fitting uses 5 arguments
+                    
+                    kernel = constantSVMoptions.kernel;
+                    
+                    if isfield(constantSVMoptions, 'kernelScale')
+                        kernelScale = constantSVMoptions.kernelScale;
+                        
+                        % then we're validating the C and not this
+                        model = fittingFunction( yTr, tXTr, kernel, kernelScale, parameter(i) );
+                        
+                    elseif isfield(constantSVMoptions, 'C')
+                        
+                        C = constantSVMoptions.C;
+                        % then we're validating the kernalScale and not this
+                        model = fittingFunction( yTr, tXTr, kernel, parameter(i), C);
+                    end
+                    
                 case 10 % only NN takes so many input arguments !
                     
                     dimensions = constantNNoptions.dimensions;
@@ -106,7 +123,7 @@ for s = 1:noOfSeeds
                         
                         dropout = constantNNoptions.dropout;
                         
-                        NN = fittingFunction( tXTr, yTr, dimensions, noEpochs,...
+                        model = fittingFunction( tXTr, yTr, dimensions, noEpochs,...
                             batchSize, 0, learningRate, dropout, parameter(i), activFun );
                         
                     elseif isfield(constantNNoptions, 'L2Weight')      
@@ -115,7 +132,7 @@ for s = 1:noOfSeeds
[…]

### 2014-12-11 Project 2/Youssef/svmKCV.m
added parallel processing to KCV

@@ -28,7 +28,7 @@ y = labels;
 
 normalizeFlag = 1;
 K = 4;
-noOfSeeds = 3;
+noOfSeeds = 2;
 constantSVMoptions.kernel = 'rbf';
 
 %% KCVating the kernel scale (sigma for rbf)
@@ -39,9 +39,13 @@ constantSVMoptions.kernel = 'rbf';
 % WHILE (test to see how much 1 model takes to train, if it's t minutes than the
 % code is going to take t*noOfSeeds*K*noOfSigmas/60 hours !! For 10 minutes
 % with the current parameters if it's 10 minutes per model it's 10*3*4*4/60
-% = 8 hours. Measure time and change parameters yourself
+% = 8 hours. Measure time and change parameters yourself. 
 
-noOfSigmas = 3;
+%This is with the old generic KCV function, the updated one using parallel
+%pools is faster, but no guarantee on how much faster, though probably a
+%lot.
+
+noOfSigmas = 12;
 sigmas = logspace(-1, 3, noOfSigmas);
 constantSVMoptions.C = []; % indicates that this isn't the parameter to genericKCV
 

### 2014-12-11 Project 2/genericKCV.m

@@ -24,14 +24,20 @@ noOfParamVals = max(1, length(parameter));
 errorsTr = zeros(noOfSeeds, K, noOfParamVals);
 errorsTe = zeros(noOfSeeds, K, noOfParamVals);
 
+parpool
+
 % K-fold cross validation
-for s = 1:noOfSeeds
+parfor s = 1:noOfSeeds
+    
+    tempTr = zeros(K, noOfParamVals);
+    tempTe = zeros(K, noOfParamVals);
     
     % split data in K fold (we will only create indices)
     setSeed(s);
     N = size(y,1);
     idx = randperm(N);
     Nk = floor(N/K);
+    idxCV = zeros(K, Nk);
     for k = 1:K
         idxCV(k,:) = idx(1+(k-1)*Nk:k*Nk);
     end
@@ -49,25 +55,27 @@ for s = 1:noOfSeeds
             
             fittingFunctionType = nargin(fittingFunction);
             
+            if normalizeFlag
             % handling normalization
             % this ASSUMES that all input variables are gaussian
             % distributed
-            switch fittingFunctionType
-                case {5, 10}
-                    % these cases correspond to the cases where inputX = tX
-                    % ( has a column of ones in the beginning
-                    
-                    [tXTr, mu, sigma] = zscore(tXTr); % train, get mu and std
-                    tXTe = normalize(tXTe, mu, sigma);  % normalize test data
-                   
-                otherwise
-                    % these cases correspond to the cases where inputX = X
-                    
-                    [normXTr, mu, sigma] = zscore(tXTr(:, 2:end)); % train, get mu and std
-                    normXTe = normalize(tXTe(:, 2:end), mu, sigma);  % normalize test data
-                    
-                    tXTr = [ones(size(normXTr, 1), 1) normXTr];
-                    tXTe = [ones(size(normXTe, 1), 1) normXTe];
+                switch fittingFunctionType
+                    case {5, 10}
+                        % these cases correspond to the cases where inputX = tX
+                        % ( has a column of ones in the beginning
+
+                        [tXTr, mu, sigma] = zscore(tXTr); % train, get mu and std
+                        tXTe = normalize(tXTe, mu, sigma);  % normalize test data
+
+                    otherwise
+                        % these cases correspond to the cases where inputX = X
+
+                        [normXTr, mu, sigma] = zscore(tXTr(:, 2:end)); % train, get mu and std
[…]

### 2014-12-12 Project 2/Youssef/svmKCV.m
updated some some + KCV bugfix + added graph

@@ -28,7 +28,7 @@ y = labels;
 
 normalizeFlag = 1;
 K = 4;
-noOfSeeds = 2;
+noOfSeeds = 6;
 constantSVMoptions.kernel = 'rbf';
 
 %% KCVating the kernel scale (sigma for rbf)
@@ -45,8 +45,9 @@ constantSVMoptions.kernel = 'rbf';
 %pools is faster, but no guarantee on how much faster, though probably a
 %lot.
 
-noOfSigmas = 12;
-sigmas = logspace(-1, 3, noOfSigmas);
+noOfSigmas = 6;
+% sigmas = logspace(-1, 3, noOfSigmas);
+sigmas = logspace(30, 200, noOfSigmas);
 constantSVMoptions.C = []; % indicates that this isn't the parameter to genericKCV
 
 [errorsTr, errorsTe] = genericKCV( y, X,...
@@ -54,4 +55,4 @@ constantSVMoptions.C = []; % indicates that this isn't the parameter to genericK
    [], constantSVMoptions, normalizeFlag);
 
 % it saves the results directly, these are to be shared
-save('SVM KCV results (light)','errorsTr', 'errorsTe', 'noOfSeeds', 'K', 'sigmas', 'constantSVMoptions');
+save('SVM KCV results (light) - focalized','errorsTr', 'errorsTe', 'noOfSeeds', 'K', 'sigmas', 'constantSVMoptions');

### 2014-12-12 Project 2/Youssef/youssefplaying2.m

@@ -1,8 +1,8 @@
 % youssef's playing with the data script, project 2
 
-% clear all;
-% close all;
-% clc
+clear all;
+close all;
+clc
 
 projectDir = '/users/youssef/Documents/Matlab/PCML/Projects/Project 2';
 addpath(genpath(projectDir));
@@ -133,9 +133,21 @@ nnPredWithReg = predictNNBinaryOutput(Te.normX, trainedNN3 );
 
 %% Training neural networks - simple NN + dropout + L2 regularization
 rand('seed',8339);
-[ trainedNN3 ] = trainNewBinaryNN(Tr.normX, Tr.y, dimensions, noEpochs, ...
+[ trainedNN4 ] = trainNewBinaryNN(Tr.normX, Tr.y, dimensions, noEpochs, ...
     batchSize, plotFlag, learningRate, dropout, [], 'sigm' );
-nnPredWithDropoutAndSigm = predictNNBinaryOutput(Te.normX, trainedNN3 );
+nnPredWithDropoutAndSigm = predictNNBinaryOutput(Te.normX, trainedNN4 );
+
+%% Training neural networks - final dropout + L2 regularization + sigm
+
+rand('seed',8339);
+
+dimensions = [size(Tr.X,2) 50 25 10 2];
+lambda = 10^-6;
+dropout = 0.3;
+
+[ trainedNN5 ] = trainNewBinaryNN(Tr.normX, Tr.y, dimensions, noEpochs, ...
+    batchSize, plotFlag, learningRate, dropout, [], 'sigm' );
+nnPredWithDropoutAndSigm = predictNNBinaryOutput(Te.normX, trainedNN5 );
 
 %% comparing results
 rand('state',8339);

### 2014-12-12 Project 2/analyzeKCVresults.m

@@ -1,5 +1,5 @@
 function [bestParameter, achievedErrorTe ] = analyzeKCVresults( errorTe,...
-    errorTr, parameterName, parameterValues, givenTitle, errorName, modeFunction )
+    errorTr, parameterName, parameterValues, givenTitle, errorName, modeFunction, logScaleFlag )
 % plots the results from cross-validating the given parameter, for
 % inspecttion and also retrieves the optimal parameter value and returns it
 
@@ -7,14 +7,26 @@ function [bestParameter, achievedErrorTe ] = analyzeKCVresults( errorTe,...
 avgErrorTe = mean(errorTe);
 avgErrorTr = mean(errorTr);
 
-figure;
-plot(parameterValues', errorTe','r-','color',[1 0.7 0.7]);
-hold on
-plot(parameterValues, errorTr','b-','color',[0.7 0.7 1]);
-plot(parameterValues, avgErrorTe' ,'r-','linewidth', 3); hold on
-plot(parameterValues, avgErrorTr','b-','linewidth', 3); xlabel(parameterName);
-ylabel(errorName)
-title(givenTitle);
+if ~logScaleFlag
+    figure;
+    plot(parameterValues', errorTe','r-','color',[1 0.7 0.7]);
+    hold on
+    plot(parameterValues, errorTr','b-','color',[0.7 0.7 1]);
+    plot(parameterValues, avgErrorTe' ,'r-','linewidth', 3); hold on
+    plot(parameterValues, avgErrorTr','b-','linewidth', 3); xlabel(parameterName);
+    ylabel(errorName)
+    title(givenTitle);
+else
+
+    figure;
+    semilogx(parameterValues', errorTe','r-','color',[1 0.7 0.7]);
+    hold on
+    semilogx(parameterValues, errorTr','b-','color',[0.7 0.7 1]);
+    semilogx(parameterValues, avgErrorTe' ,'r-','linewidth', 3); hold on
+    semilogx(parameterValues, avgErrorTr','b-','linewidth', 3); xlabel(parameterName);
+    ylabel(errorName)
+    title(givenTitle);
+end
 
 % computes the best parameter (usually a penalizazion term lambda) that
 % achieves the best test Error, returns both

### 2014-12-12 Project 2/genericKCV.m

@@ -24,8 +24,6 @@ noOfParamVals = max(1, length(parameter));
 errorsTr = zeros(noOfSeeds, K, noOfParamVals);
 errorsTe = zeros(noOfSeeds, K, noOfParamVals);
 
-parpool
-
 % K-fold cross validation
 parfor s = 1:noOfSeeds
     
@@ -176,7 +174,7 @@ parfor s = 1:noOfSeeds
     end
     
     errorsTr(s, :, :) = tempTr;
-    errorsTr(s, :, :) = tempTe;
+    errorsTe(s, :, :) = tempTe;
 end
 
 % averages over the k folds

### 2014-12-13 Project 2/kevin/mr_mean.m
mr_mean predictions Don't understand why pred = pairs instead of pred = mean

@@ -1,15 +1,26 @@
+%% initialize 
+clear all;
+load songTrain;
+load songTestPairs;
+Ytest_weak_pred = Ytest_weak_pairs;
+Ytest_strong_pred = Ytest_strong_pairs;
+percent = 10;
+i = 1;
+%% 
 
-splitData;
-%%
-
+[Ytrain_new]=splitDataMRmean(Ytrain,Gtrain,percent/100);
+    
 m = computeMeanPerArtist(Ytrain_new);
 % repmat on m
-
-repm = repmat(m,size(Ytrain_new,1),1);
-
-% Don't take into account the 0 count
-idx = find(Ytest_weak >0);
-
+m(m==0) = 1;
+repm_weak = repmat(m,size(Ytest_weak_pairs,1),1);
+repm_strong = repmat(m,size(Ytest_strong_pairs,1),1);
+idxw = find(Ytest_weak_pairs>0);
+idxs = find(Ytest_strong_pairs>0);
+Ytest_weak_pred(idxw) = repm_weak(idxw);
+Ytest_strong_pred(idxs) = repm_strong(idxs);
+rmse = sqrt(2*computeCost(Ytest_weak_pairs,Ytest_weak_pred))    
 %%
-rmseTr = rmse(Ytest_weak(idx),repm(idx))
-%rmseTr = sqrt(2*computeCost(Ytest_weak,repm))
+
+save('songPred', 'Ytest_strong_pred', 'Ytest_weak_pred');
+

### 2014-12-13 Project 2/kevin/mr_similarity.m

@@ -6,7 +6,7 @@ splitData;
 ua_tr = Ytrain_new; % users-artists 
 ua_te = Ytest_weak;
 S = artistName;
-
+%%
 
 rmse = wrapperCF(ua_tr,ua_te,S);
 

### 2014-12-13 Project 2/kevin/splitDataMRmean.m

@@ -1,0 +1,45 @@
+function [Ytrain_new ] = splitDataMRmean(Ytrain,Gtrain,testPercent)
+% The following script creates two kinds of test sets
+% Ytest_strong contains pairs for new users to test (the non-zero entries).
+% Ytest_weak contains pairs for existing users to test (the non-zero entries).
+% Gstrong contains the friendship information of new users with all users
+% Ytrain_new and Gtrain_new is the new training set 
+
+% Test data for Strong generalization
+% keep 10% of users for testing as 'new users'
+% You should decide on your own how many new users you want to test on
+setSeed(1);
+Ytrain(Ytrain~=0) = log(Ytrain(Ytrain~=0));
+nU = size(Ytrain,1);
+idx = randperm(nU);
+nTe = floor(nU*testPercent); 
+idxTe = idx(1:nTe);
+idxTr = idx(nTe+1:end);
+Ytrain_new = Ytrain(idxTr,:);
+Ytest_strong = Ytrain(idxTe,:);
+Gtrain_new = Gtrain(idxTr, idxTr);
+Gstrong = Gtrain(idxTe, [idxTr idxTe]);
+
+% Test data for weak generalization
+% Keep 10 entries per existing user as test data
+[D N] = size(Ytrain_new);
+numD = 10; % number of artists held out per user
+dd = [];
+nn = [];
+yy = [];
+for n = 1:N
+    On = find(Ytrain_new(:,n)~=0);
+    if length(On)>2
+        ind = unidrnd(length(On),numD,1); % choose some for testing
+        d = On(ind);
+        dd = [dd; d];
+        nn = [nn; n*ones(numD,1)];
+        yy = [yy; Ytrain_new(d,n)];
+    end
+end
+Ytest_weak = sparse(dd,nn,yy,D,N);
+Ytrain_new(sub2ind([D N], dd, nn)) = 0;
+
+
+end
+

### 2014-12-13 Project 2/kevin/wrapperCF.m

@@ -4,9 +4,26 @@ function predictX = wrapperCF( ua_tr, ua_te, S )
 % ua_tr_tr : Ytrain_train
 % ua_tr_te : Ytrain_test 
 % S : Artists_name
-%%
-
-
+%% top-k count for u users
+k = 5;
+u = [1;109;472;788;982];
+l = length(u);
+topKL = cell(l+1,k+1);
+for j = 2:k+1
+    topKL(1,j)={j-1};
+end
+topKL_rec = topKL;
+for i = 1:l
+    [sortX,sortIdx] = sort(ua_tr(u(i),:),'descend'); % sort the scores for user bob
+    maxKVal = sortX(1:k);
+    maxKindx = sortIdx(1:k);
+    topKArtists = S(maxKindx); % get the name of the artists
+    topKL(i+1,1) = {i};
+    topKL(i+1,2:end) = topKArtists;
+end
+topKL
+fprintf('Top-k for l selected users. Ready to process with recommendation. It can take some time.\n Press to continue...');
+pause;
 
 %% number of listening in average per user
 users_mean = computeMeanPerUser(ua_tr);
@@ -64,7 +81,7 @@ fullSu = full(Su);
 X = sparse(Su * ua_tr);
 
 %% Get the score for each 0
-keyboard
+
 umrep = repmat(users_mean,1,size(X,2));
 X = X';
 ua_col= X(:);
@@ -81,28 +98,24 @@ predictX = ua_col.*umrep;
 ua_te = ua_te';
 ua_tecol = ua_te(:);
 %ua_tecol = ua_tecol(idx0);
-keyboard
-rmse = computeCost(ua_tecol,predictX)
+
+rmse = computeCost(ua_tecol,predictX);
 
 
 %% get the top-k values and their indices
-k = 5;
-l = 5;
-% Get the top k recommendation for the l first users
-topKL = cell(l+1,k+1);
-for j = 2:k+1
-    topKL(1,j)={j-1};
-end
+% Get the top k recommendation for the l users specified at the begining
+
 for i = 1:l
     % check if we remove the non-predicted values
-    [sortX,sortIdx] = sort(X(i,:),'descend'); % sort the scores for user bob
+    [sortX,sortIdx] = sort(X(u(i),:),'descend'); % sort the scores for user bob
     maxKVal = sortX(1:k);
     maxKindx = sortIdx(1:k);
[…]

### 2014-12-13 Project 2/kevin/mr_mean.m
mr mean bug idx

@@ -21,6 +21,4 @@ Ytest_weak_pred(idxw) = repm_weak(idxw);
 Ytest_strong_pred(idxs) = repm_strong(idxs);
 rmse = sqrt(2*computeCost(Ytest_weak_pairs,Ytest_weak_pred))    
 %%
-
 save('songPred', 'Ytest_strong_pred', 'Ytest_weak_pred');
-

### 2014-12-13 Project 2/Youssef/youssefplaying2.m
modifications to kevin

@@ -1,8 +1,8 @@
 % youssef's playing with the data script, project 2
 
-clear all;
-close all;
-clc
+% clear all;
+% close all;
+% clc
 
 projectDir = '/users/youssef/Documents/Matlab/PCML/Projects/Project 2';
 addpath(genpath(projectDir));
@@ -141,7 +141,7 @@ nnPredWithDropoutAndSigm = predictNNBinaryOutput(Te.normX, trainedNN4 );
 
 rand('seed',8339);
 
-dimensions = [size(Tr.X,2) 50 25 10 2];
+dimensions = [size(Tr.X,2) 250 50 10 2];
 lambda = 10^-6;
 dropout = 0.3;
 

### 2014-12-13 Project 2/genericKCV.m

@@ -124,13 +124,28 @@ parfor s = 1:noOfSeeds
                     dimensions = [size(tXTr, 2) dimensions];
                     
                     if isfield(constantNNoptions, 'dropout')
-                        % we're varying the L2 weight and keeping the
-                        % dropout constant in this case
                         
-                        dropout = constantNNoptions.dropout;
-                        
-                        model = fittingFunction( tXTr, yTr, dimensions, noEpochs,...
-                            batchSize, 0, learningRate, dropout, parameter(i), activFun );
+                        if isfield(constantNNoptions, 'L2Weight')
+                            % both L2 weight and dropout are fixed, then
+                            % changing the dimensions of the NN
+                            
+                            dropout = constantNNoptions.dropout;
+                            L2Weight = constantNNoptions.L2Weight;
+                            
+                            dimensions = [size(tXTr, 2) parameter{i}];
+                            
+                            model = fittingFunction( tXTr, yTr, dimensions, noEpochs,...
+                                batchSize, 0, learningRate, dropout, L2Weight, activFun );
+                        else
+
+                            % we're varying the L2 weight and keeping the
+                            % dropout constant in this case
+
+                            dropout = constantNNoptions.dropout;
+
+                            model = fittingFunction( tXTr, yTr, dimensions, noEpochs,...
+                                batchSize, 0, learningRate, dropout, parameter(i), activFun );
+                        end
                         
                     elseif isfield(constantNNoptions, 'L2Weight')      
                         % we're varying the dropout and keeping the

### 2014-12-13 Project 2/kevin/mr_mean.m

@@ -2,8 +2,18 @@
 clear all;
 load songTrain;
 load songTestPairs;
-Ytest_weak_pred = Ytest_weak_pairs;
-Ytest_strong_pred = Ytest_strong_pairs;
+Ytest_weak_pred = full(Ytest_weak_pairs);
+Ytest_strong_pred = full(Ytest_strong_pairs);
+
+Ytest_weak_pairs = full(Ytest_weak_pairs);
+Ytest_strong_pairs = full(Ytest_strong_pairs);
+
+
+Ytrain = full(Ytrain);
+Gtrain = full(Gtrain);
+
+Ytest = Ytrain(Ytest_weak_pairs);
+
 percent = 10;
 i = 1;
 %% 
@@ -13,12 +23,15 @@ i = 1;
 m = computeMeanPerArtist(Ytrain_new);
 % repmat on m
 m(m==0) = 1;
-repm_weak = repmat(m,size(Ytest_weak_pairs,1),1);
-repm_strong = repmat(m,size(Ytest_strong_pairs,1),1);
-idxw = find(Ytest_weak_pairs>0);
-idxs = find(Ytest_strong_pairs>0);
-Ytest_weak_pred(idxw) = repm_weak(idxw);
-Ytest_strong_pred(idxs) = repm_strong(idxs);
-rmse = sqrt(2*computeCost(Ytest_weak_pairs,Ytest_weak_pred))    
+repm_weak = full(repmat(m,size(Ytest_weak_pairs,1),1));
+repm_strong = full(repmat(m,size(Ytest_strong_pairs,1),1));
+idxw = Ytest_weak_pairs>0;
+idxs = Ytest_strong_pairs>0;
+repm_weak = exp(repm_weak);
+repm_weak(~idxw) = 0;
+
+% Ytest_strong_pred(idxs) = repm_strong(idxs);
+
+rmse = rmse(Ytest_weak_pred, Ytest_weak_pairs);
 %%
 save('songPred', 'Ytest_strong_pred', 'Ytest_weak_pred');

### 2014-12-13 Project 2/rmse.m

@@ -1,7 +1,7 @@
 function [ result ] = rmse( x, y )
 % computes the RMSE between two data vectors
 
-result = sqrt(mean((x(:)-y(:)).^2));
+result = sqrt(nanmean((x(:)-y(:)).^2));
 
 end
 

### 2014-12-13 Project 2/report/sampleReport/paper.tex
report update

\documentclass{article} % For LaTeX2e
% We will use NIPS submission format
\usepackage{nips13submit_e,times}
% for hyperlinks
\usepackage{hyperref}
\usepackage{url}
% For figures
\usepackage{graphicx} 
\usepackage{subfigure} 
% math packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsopn}
\usepackage{ifthen}
\usepackage{natbib}

\title{Project-I by Group CITYNAME}

\author{
Mohammad Emtiyaz Khan\\
EPFL \\
\texttt{emtiyaz@gmail.com} \\
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\nipsfinalcopy

\begin{document}

\maketitle

\begin{abstract}
In this report, we summarize our findings for the demo done on October 23, 2014.
We fit several linear methods to the regression dataset and compare them.
We found that the input matrix is ill-conditioned and therefore ridge regression is prefered over least-squares. It also appears that the linear model is appropriate for this data and predicts the data reasonably well. We also investigated a few feature transformations but found no significant improvements.
\end{abstract}

\section{Data Description}
Our train-data consists of output variables $\mathbf{y}$ and input variables $\mathbf{X}$. We have $N=600$ data examples. Each input vector $\mathbf{x}_n$ is of dimensionality $D=10$. Out of these 10 variables, 8 variables are real valued while 1 variable is binary and 1 variable is categorical with 4 categories.

We also have test-data where we do not observe $\mathbf{y}$. We have $N=600$ test examples. Our goal is to produce predictions for test examples, as well as an approximation of the test-error.

\section{Data visualization and cleaning}
We performed basic exploratory data analysis on our data. Figure \ref{fig:boxplotX} shows distributions of all 10 input variables. As expected the data is not centered, and therefore we normalize the data. Figure \ref{fig:histY} shows the histogram of the output variable. We see that there are two data points which have rather high value of the output variables compared to rest of the data. Since these are only two data points, we remove them from the data (example number 362 and 402).

We also investigated the correlation between output and input variables (plot not shown, since it is not important for our analysis). We found that most of the variables are correlated with the output, confirming the need for a full regression analysis. The third variable did not seem significantly correlated, however this preliminary observation does not suggest removal of this variable from the dataset.

We use a dummy encoding for the categorical variable. The binary variable does not require any dummy encoding. This gives us a total of 14 input variables.

The input matrix $\mathbf{X}$ is rank-deficient with a rank of 6, instead of 10. This implies that we must ``lift" the eigenvalues, using methods such ridge regression.

\begin{figure}[!t]
\center
\subfigure[Boxplot of real-valued $\mathbf{X}$. Data is not centered and therefore we normalize it.]{\includegraphics[width=2.5in]{figures/boxplotX.pdf} \label{fig:boxplotX}}
\hfill
\subfigure[Histogram of $\mathbf{y}$. We can clearly see two outliers.]{\includegraphics[width=2.5in]{figures/histY.pdf} \label{fig:histY}}
\caption{}
\end{figure}

\section{Ridge regression}
We applied least-squares and ridge regression to this dataset. Since the matrix is ill-conditioned, least-squares is not suitable. Therefore, we report results obtained with ridge regression alone. Note that the improvements using ridge regression were modest and not much lower than that of linear regression, however we do not expect least-squares to work well when there is a lot of testing data available.

Figure \ref{fig:ridgeCurve} shows the results obtained with ridge regression when we use $50\%$ of the data as test data and rest as training data. We varied the value of $\lambda$ from $10^{-4}$ to $10^3$, choosing total 500 points in between. We can see that there is a small improvement obtained for some values of $\lambda$.

We did experiments to plot a learning curve for this data (see Andrew Ng's notes about the learning curve). We held out 20\% of data as test data and rest as training data. We chose to slowly increase the proportion of data used for training. For each proportion of the training data, we repeated the experiment 30 times to compute the distribution of error. We fit ridge regression to each sampled training set and test it on the same $20\%$ test data. We varied the value of $\lambda$ from $10^{-4}$ to $10^3$, choosing total 500 points in between.

This gives us the learning curve shown in Fig. \ref{fig:learnCurve}. The blue curve shows the train error while red curve shows the test error. We can see that both training and test error converge, with the variance of estimates decreasing as we increase the training data size. There is also a very small gap between the train and test errors, showing that the linear model is a reasonable choice. The small gap exists perhaps because we have only limited test data.

\section{Feature transformations}
We tried several feature transformations. We found that we get a small improvement in performance when we take $\sqrt{|X_{ni}|}$ for all entries of $\mathbf{X}$. We did not check (due to lack of time) whether it matters if we apply this to one variable or all. We performed experiments similar to the last section (although one should really do cross-validation). Values of lambda were kept same as the last section.

We compare three methods. First is a baseline where we do not use any input variables i.e. mean value of the output. The second method is the ridge regression described in previous section. The third method is ridge regression with a feature transformation. The first method gave RMSE of around 3 which was way worse than the other two methods.

The RMSE for the last two methods are shown in Fig. \ref{fig:summary}. We see that both test and train error decrease, however it appears that the improvement is very little and may not be significant.

\begin{figure}[!h]
\center
\subfigure[Ridge regression for a 50-50 split.]{\includegraphics[width=2.5in]{figures/ridgeCurve.pdf} \label{fig:ridgeCurve}}
\hfill
\subfigure[Comparison of ridge regression with and without feature transformation. The improvement is very little and might be insignificant.]{\includegraphics[width=2.7in]{figures/summary.pdf} \label{fig:summary}}
\subfigure[Learning curve. Blue is training data and red is test data.]{\includegraphics[width=4in]{figures/learningCurve.pdf} \label{fig:learnCurve}}
\caption{}
\end{figure}

\section{Summary}
In this report, we analyzed a regression dataset and found that ridge regression is a reasonable fit. We estimate that the average test error is 1.213 ($\pm$ 0.02). We tried some feature transformation and found that there is a small improvement giving us a test error of around 1.198 ($\pm$ 0.015). This improvement, however, is not significant.

\subsubsection*{Acknowledgments}
We would like to thank Timur for making the dataset for this study, Carlos and Ilija to help conduct the experiments, and all the students who attended the session making it exciting for us. All the code, as well as this report, was written by Emti.

\subsubsection*{References}

\end{document}

### 2014-12-13 Project 2/Person detection example/personDetectionExample.m

@@ -71,7 +71,7 @@ addpath(genpath('where/the/deeplearningtoolboxis/'));
 %
 % Make sure you understand what is going on below!
 
-rand('state',8339);  % fix seed, this    NN is very sensitive to initialization
+rng('default');  % fix seed, this    NN is very sensitive to initialization
 
 % setup NN. The first layer needs to have number of features neurons,
 %  and the last layer the number of classes (here two).

### 2014-12-13 Project 2/kevin/wrapperCF.m

@@ -75,7 +75,7 @@ Su = Sup1*invSqrtP;
 %Si = invSqrtQ*RtR*invSqrtQ;
 %Sip1 = bsxfun(@times,diag(invSqrtQ),RtR);
 %Si = Sip1*invSqrtQ;
-fullSu = full(Su);
+
 
 %% user-user clollaborative (part 3.a)
 X = sparse(Su * ua_tr);
@@ -83,6 +83,7 @@ X = sparse(Su * ua_tr);
 %% Get the score for each 0
 
 umrep = repmat(users_mean,1,size(X,2));
+ummax = repmat(max(ua_tr),1,size(X,2));
 X = X';
 ua_col= X(:);
 %idx0 = find(ua_col>0);
@@ -90,7 +91,8 @@ ua_col= X(:);
 umrep = umrep(:);
 %umrep = umrep(idx0);
 
-predictX = ua_col.*umrep;
+predictX = (ua_col.*ummax).^0.8;
+
 %for i = 1:size(X0,1)
  %   predictX(i,:) = X0(i,:).*users_mean(i);
 %end
@@ -99,8 +101,8 @@ ua_te = ua_te';
 ua_tecol = ua_te(:);
 %ua_tecol = ua_tecol(idx0);
 
-rmse = computeCost(ua_tecol,predictX);
-
+rmse = computeCost(ua_tecol,predictX)
+pause;
 
 %% get the top-k values and their indices
 % Get the top k recommendation for the l users specified at the begining

### 2014-12-13 Project 2/splitData.m
splitdata

@@ -1,0 +1,43 @@
+% The following script creates two kinds of test sets
+% Ytest_strong contains pairs for new users to test (the non-zero entries).
+% Ytest_weak contains pairs for existing users to test (the non-zero entries).
+% Gstrong contains the friendship information of new users with all users
+% Ytrain_new and Gtrain_new is the new training set 
+clear all;
+load songTrain;
+
+% Test data for Strong generalization
+% keep 10% of users for testing as 'new users'
+% You should decide on your own how many new users you want to test on
+setSeed(1);
+Ytrain(Ytrain~=0) = log(Ytrain(Ytrain~=0));
+nU = size(Ytrain,1);
+idx = randperm(nU);
+nTe = floor(nU*0.1); 
+idxTe = idx(1:nTe);
+idxTr = idx(nTe+1:end);
+Ytrain_new = Ytrain(idxTr,:);
+Ytest_strong = Ytrain(idxTe,:);
+Gtrain_new = Gtrain(idxTr, idxTr);
+Gstrong = Gtrain(idxTe, [idxTr idxTe]);
+
+% Test data for weak generalization
+% Keep 10 entries per existing user as test data
+[D N] = size(Ytrain_new);
+numD = 10; % number of artists held out per user
+dd = [];
+nn = [];
+yy = [];
+for n = 1:N
+    On = find(Ytrain_new(:,n)~=0);
+    if length(On)>2
+        ind = unidrnd(length(On),numD,1); % choose some for testing
+        d = On(ind);
+        dd = [dd; d];
+        nn = [nn; n*ones(numD,1)];
+        yy = [yy; Ytrain_new(d,n)];
+    end
+end
+Ytest_weak = sparse(dd,nn,yy,D,N);
+Ytrain_new(sub2ind([D N], dd, nn)) = 0;
+

### 2014-12-13 Project 2/Youssef/youssefplaying2.m
added PD TPR comparison plot

@@ -143,7 +143,7 @@ rand('seed',8339);
 
 dimensions = [size(Tr.X,2) 250 50 10 2];
 lambda = 10^-6;
-dropout = 0.3;
+dropout = 0.43;
 
 [ trainedNN5 ] = trainNewBinaryNN(Tr.normX, Tr.y, dimensions, noEpochs, ...
     batchSize, plotFlag, learningRate, dropout, [], 'sigm' );
@@ -157,6 +157,19 @@ randPred = rand(size(Te.y))*2-1;
 % and plot all together, and get the performance of each method
 
 % this is to show it in the legend
-methodNames = {'NN with dropout = 0.3 and sigmoid', 'NN with L2 reg = 1/1000', 'NN with dropout = 0.3', 'NN with sigmoid activ.' 'Neural Network (NN)', 'Random'};
+methodNames = {'NN with dropout = 0.43 and sigmoid', 'NN with L2 reg = 1/1000', 'NN with dropout = 0.3', 'NN with sigmoid activ.' 'Neural Network (NN)', 'Random'};
 
 avgTPRList = evaluateMultipleMethods( Te.y > 0, [nnPredWithDropoutAndSigm, nnPredWithReg, nnPredWithDropout, nnPredSigm, nnPred,randPred], true, methodNames );
+
+%% comparing results - with PLR & SVD in workspace
+
+rand('state',8339);
+% let's also see how random predicition does
+randPred = rand(size(Te.y))*2-1;
+
+% and plot all together, and get the performance of each method
+
+% this is to show it in the legend
+methodNames = {'SVM with RBF, sigma = 81', 'PLR with lambda = 44', 'NN with L2 weight = 10^6 dropout = 0.43 and sigmoid', 'NN with L2 reg = 1/1000', 'NN with dropout = 0.43', 'NN with sigmoid activ.' 'Neural Network (NN)', 'Random'};
+
+avgTPRList = evaluateMultipleMethods( Te.y > 0, [SVMSoftpredictions, PLRPred, nnPredWithDropoutAndSigm, nnPredWithReg, nnPredWithDropout, nnPredSigm, nnPred,randPred], true, methodNames );

### 2014-12-13 Project 2/penLogisticRegression.m

@@ -8,6 +8,6 @@ maxIters = 3000;
 betaStart = zeros(size(tX, 2), 1);
 
 beta = gradientDescent(y, tX, maxIters, alpha, betaStart,...
-    @computeCostLogReg, @computeGradientLogReg, lambda, 0);
+    @computeCostLogReg, @computeGradientLogReg, lambda, 1);
 
 end

### 2014-12-13 Project 2/report/paper.tex

@@
 
+We cross-validated parameters for the NN: namely the dropout, $L_2$ weight and dimension (depth mostly).
+Following are the corresponding plots, respectively.
+We can see that a dropout value of 0.43 is actually the best one to choose (the one with minimal test error, red curve) This was done for 15 values of the dropout, $k=4$, and 15 different seeds.
+We then cross-validated the $L_2$ weight penalty parameter for NN and obtained an optimal TPR of 81.45\% for a weight penalty of $10^{-6}$ (while setting the dropout to a static value of 0.43).
+We used 15 seeds and 20 values of the $L_2$ weight, with $k = 4$.
+\begin{figure}
+\centering     %%% not \center
+\subfigure[NN with various dropouts]{\label{fig:histC}\includegraphics[width=0.49\textwidth]{images/NNKCVplotdropout.png}}
+\subfigure[NN with various L2 weight penalties]{\label{fig:loghist}\includegraphics[width=0.49\textwidth]{images/NNKCVplotL2weight.png}}
+\caption{Test errors}
+\end{figure}
+
 \subsection{Penalize Logistic regression (cross-validation for $\lambda$}
@@
 That decision is very difficult to make, especially considering that music listening - as any human-  behaviour cannot be expected to always follow mathematical rules.
-\textcolor{red}{What about poisson distribution for count?}
 
@@
 Then we applied the log transformation to obtain a more Gaussian input as seen in Figure \ref{fig:loghist}.
-We then compute the RMSE based on the same principle as before and rescale the result with the exponential to obtain an RMSE of 4130, which is already much better.
+We then compute the RMSE based on the same principle as before and rescale the result with the exponential to obtain an RMSE of 4130 (8.3 in log scale), which is already much better.
 
 \subsection{Collaborative-filtering}
+Let $UA$ (users-artists) be equal to $Y_train$.
+The values inside the matrix are the number of time an user heard an artist.
+One of the principle features of the problem now is to say whether a user “likes” an artist of not, and ideally to know how much.
 
+We first tried to normalize each row independently so that we will get values between 0 and 1.
+Then we applied the following method : We created the matrix $P$, which is a diagonal matrix where each value in the diagonal is the total listening count for each user, (there are problems with users with total count $= 0$, however those don’t exist in our dataset).
+Hence, this matrix is invertible.
+
+Using this inverse we can compute the similarity between users.
+At this point we don’t take into account the social graph, but it looks promising to refer later (if we can assume that friends are more likely to like the same artists).
+The similarity matrix X should contain 1 in the diagonal ( a user is maximally similar to himself).
+However this is not the case in our dataset : when we computed the similarity of the user using the formula $UA\cdot UA^T$ then we obtain the total number of listening in common with other users.
+That means to get a diagonal of $1$’s, the matrix $P$ should be defined differently : instead of having a diagonal matrix where each entry is  total listening count of user $i$, we should simply have the same values as $UA \cdot UA^T$ in the diagonal matrix $P$ and $0$ elsewhere.
+Then we will measure the cosine similarity between another user.
+Note that this is possible to do without normalizing the data, as we obtain the same results regardless.
+
+One idea is to use the social graph along with the similarity matrix we obtained is to see if it is indeed the case that friends are similar users ( listeners), but not the other way around.
+Therefore, when we want to add a new users and predict his counts (taste in music), one possibility would be to use the information in the social graph and then use the similarity matrix to predict.
+
+So finding the social relation between users requires finding the position of $1$s in the social graph.
+With these positions we can check what is the similarity between friends and compute the average similarity.
+We obtained experimentally an average of $0.2145$, which means their taste are not as correlated as we thought, putting the usefulness of the social graph into question again.
+We can use this number when we want to predict counts for a new user: we can look into the social graph to obtain the relationship between the new user and all existing users and then decide on their similarity between them.
+To predict counts we can sum the counts for the friend's user for a particular artist and divide it by the number of his friends.
 \section{Summary}

### 2014-12-13 Project 2/Youssef/dimensions_generator.m
added my mat files

@@ -1,0 +1,10 @@
+for i = 1:length(depths)
+    temp = zeros(1, depths(i));
+    temp(1) = 2;
+    
+    for j = 2:depths(i)
+        temp(j) = temp(j-1)*4;
+    end
+    
+    dimensions{i} = temp(end:-1:1);
+end

### 2014-12-14 Project 2/report/paper.tex
paper final to review

@@
 \begin{abstract}
-
+In this project we are given two problems: that of predicting the music listening counts for a given set of users based on their previous listening behaviour and friendship relations; and that of recognizing the presence of people no images.
+This report discusses preliminary observations on the given training data, and then results of testing various ML methods for prediction / classification.
 \end{abstract}
@@
 We used 15 seeds and 20 values of the $L_2$ weight, with $k = 4$.
+
+A last thing we wanted to cross validate was the depth, but due to time restrictions and a huge computational workload for doing so (for a high number of units in profound NNs), we simply decided to pick dimensions based on a try out of various ones by hand.
 \begin{figure}
 \centering     %%% not \center
-\subfigure[NN with various dropouts]{\label{fig:histC}\includegraphics[width=0.49\textwidth]{images/NNKCVplotdropout.png}}
-\subfigure[NN with various L2 weight penalties]{\label{fig:loghist}\includegraphics[width=0.49\textwidth]{images/NNKCVplotL2weight.png}}
+\subfigure[NN with various dropouts]{\label{fig:nnkcvdrop}\includegraphics[width=0.49\textwidth]{images/NNKCVplotdropout.png}}
+\subfigure[NN with various L2 weight penalties]{\label{fig:nnkcvl2}\includegraphics[width=0.49\textwidth]{images/NNKCVplotL2weight.png}}
 \caption{Test errors}
@@
 
+We then launched several small cross-validation tests and determined that the best SVM variant was the one using the RBF kernel and a $\sigma$ of 81 (plots not shown here for brevity)
+
+A comparison plot for fits using the staff-provided data split and seed, using the various methods, is given in Figure \ref{fig:pdplot}, using the same seed.
+It gives the test error equivalent (TPR) for each one.
 \begin{figure}
 \centering
-\includegraphics[width=1\textwidth]{images/TPRplot.png}
+\includegraphics[width=1\textwidth]{images/PDTPRcomparisonplots.png}
 \caption{Comparison of different methods for People detection}
@@
 Let $UA$ (users-artists) be equal to $Y_train$.
+We split the data into a training set $ua_{tr}$ and a test set $ua_{te}$.
 The values inside the matrix are the number of time an user heard an artist.
@@
 
-Using this inverse we can compute the similarity between users.
+Using this inverse we can compute the similarity matrix $S_u$ between users.
 At this point we don’t take into account the social graph, but it looks promising to refer later (if we can assume that friends are more likely to like the same artists).
-The similarity matrix X should contain 1 in the diagonal ( a user is maximally similar to himself).
-However this is not the case in our dataset : when we computed the similarity of the user using the formula $UA\cdot UA^T$ then we obtain the total number of listening in common with other users.
-That means to get a diagonal of $1$’s, the matrix $P$ should be defined differently : instead of having a diagonal matrix where each entry is  total listening count of user $i$, we should simply have the same values as $UA \cdot UA^T$ in the diagonal matrix $P$ and $0$ elsewhere.
+The similarity matrix $S_u$ should contain 1 in the diagonal ( a user is maximally similar to himself).
+However this is not the case in our dataset : when we computed the similarity of the user using the formula $ua_{tr}\cdot ua_{tr}^T$ then we obtain the total number of listening in common with other users.
+That means to get a diagonal of $1$’s, the matrix $P$ should be defined differently : instead of having a diagonal matrix where each entry is  total listening count of user $i$, we should simply have the same values as $ua_{tr} \cdot ua_{tr}^T$ in the diagonal matrix $P$ and $0$ elsewhere.
 Then we will measure the cosine similarity between another user.
 Note that this is possible to do without normalizing the data, as we obtain the same results regardless.
+Computing the average of listing by users $u_a$ and multiply this with the scores obtained from the similarity matrix ($S_c = S_u \cdot ua_{tr}$).
+After playing with numbers, we discovered that we can predict the counts $P_r = (X\cdot u_a)^{0.8}$ where $P_r$ is the predictions.
+When we compare this prediction with the test set $ua_{te}$, we obtain a rmse of $3.06$ in log scale, which is quite accurate.
+This is the weak prediction.
 
 One idea is to use the social graph along with the similarity matrix we obtained is to see if it is indeed the case that friends are similar users ( listeners), but not the other way around.
-Therefore, when we want to add a new users and predict his counts (taste in music), one possibility would be to use the information in the social graph and then use the similarity matrix to predict.
+Therefore, when we want to add a new users and predict his counts (taste in music), one possibility would be to use the information in the social graph and then use the similarity matrix to predict
 
@@
 We can use this number when we want to predict counts for a new user: we can look into the social graph to obtain the relationship between the new user and all existing users and then decide on their similarity between them.
-To predict counts we can sum the counts for the friend's user for a particular artist and divide it by the number of his friends.
+This similarity for new user $i$ is computed as the mean similarity of all his friends in the known set ($Y_{train}$) for a particular user $j$.
+We obtain a new similarity matrix and can compute the predictions.
+Basically this is the same idea as before but only using friends of the new users.
+\subsubsection{Recommendations}
+Using the similarity matrix $S_u$ and the dataset, we computed $S_c = S_u \cdot UA$ as before to obtained the scores.
+With these scores we can recommend artists to existing users.
+For example in Figure \ref{fig:rec}, using the first user in the dataset, we see his top-5 listening and the top-5 recommendations.
+The method to compute this recommendation is called user-user collaborative-filtering.
+\begin{figure}[h!]
+\centering     %%% not \center
+\subfigure[Top-5]{\begin{tabular}{c|c|}
+ & User 1\\
+ \hline
+1 & Morrissey\\
+2 & The Smiths\\
+3 & Elvis Presley\\
+4 & The Housemartins \\
+5 & Blur\\
+\hline
+\end{tabular}}
+\subfigure[Top-5 recommendations]{\begin{tabular}{c|c|}
+ & User 1\\
+ \hline
+1 & Just Surrender\\
+2 & Huun-Huur-Tu\\
+3 & Beverly Kenney\\
+4 & Imperative Reaction \\
+5 & Maze featuring Frankie Beverly\\
+\hline
+\end{tabular}}
+
+\caption{Top-5 recommendation for user 1}
+\label{fig:rec}
+\end{figure}
+
 \section{Summary}
-
+For people detection, we decided to opt for NN with the cross-validated parameters instead of the otherwise best-TPR simple SVM method.
+That is because we found the output of NN predictions to be more stable than that of SVM (score between -1 and 1, in contrast to usually off-interval output for SVM).
+The final parameters used are $L_2$ weight = $10^{-6}$, a dropout of 0.43 and dimensions of $[D 250 50 10 2]$, and using sigmood internally, we also increased the number of epochs to 40.
 \subsubsection*{Acknowledgments}
-
+We would like to thanks the teaching team for their availability during the exercise sessions and office hours but also on the forum.
+We also want to thanks Youssef who organized the milestone of the project.
 \subsubsection*{References}

### 2014-12-14 Project 2/kevin/mr_mean.m
shity code

@@ -32,6 +32,5 @@ repm_weak(~idxw) = 0;
 
 % Ytest_strong_pred(idxs) = repm_strong(idxs);
 
-rmse = rmse(Ytest_weak_pred, Ytest_weak_pairs);
 %%
 save('songPred', 'Ytest_strong_pred', 'Ytest_weak_pred');

### 2014-12-14 Project 2/kevin/mr_similarity.m

@@ -2,16 +2,60 @@
 close all; clear all;
 %load('songTrain.mat');
 splitData;
+load('songTestPairs');
 %Ytrain(Ytrain>0) = 1;
 ua_tr = Ytrain_new; % users-artists 
 ua_te = Ytest_weak;
+ua = Ytrain;
 S = artistName;
+
 %%
 
-prediction = wrapperCF(ua_tr,ua_te,S);
+[predictionWeak,Su_train] = wrapperCF(ua,ua,S);
+%%
+%[dummy,Su_train] = wrapperCF(ua,ua,S);
 
 %% Are the friends similar?
-idx1 = find(Gtrain==1);
-simOfFriends = Su(idx1);
+idx1= find(Gtrain==1);
+simOfFriends = Su_train(idx1);
 meanSimilarity = mean(simOfFriends)
+
+%%  create the new similarity matrix for new users based on friends
+% For position (i,j):  user i will have similarity value mean(Similarity of
+% his friends in the train set with user j).
+Su_new = zeros(size(Gstrong,1),size(ua,1));
+
+for i = 1:size(Gstrong,1)
+    for j = 1:size(ua,1)
+        frIdx = find(Gtrain(i,:)==1);
+        simOfFriendsForUj = Su_train(frIdx,j);
+        Su_new(i,j) = mean(simOfFriendsForUj);
+    end
+end
+Su_new = sparse(Su_new);  
+%% create the final similarity matrix to compute the prediction as before
+part1 = [Su_train; Su_new];
+part2 = [Su_new';eye(size(Su_new,1))];
+Su_final = [part1 part2];
+%% Compute the strong predictions
+% ua_new is the new user-artsts matrix of size 1867x15082, but with 0 only
+% in the 93 last rows since we don't have any count for them
+ua_new = [ua;zeros(size(Su_new,1),size(ua,2))];
+strongX = sparse(Su_final*ua_new);
+PredStrong = strongX(end-92:end,:);
+
+%% Store resultsla fprintf('storing results')
+fprintf('storing results');
+idxw = find(Ytest_weak_pairs>0);
+idxs = find(Ytest_strong_pairs>0);
+Ytest_weak_pred =Ytest_weak_pairs;
+Ytest_strong_pred = Ytest_strong_pairs;
+p = predictionWeak(idxw);
+p(p==0)=1;
+Ytest_weak_pred(idxw) = p;
[…]

### 2014-12-14 Project 2/kevin/wrapperCF.m

@@ -1,4 +1,4 @@
-function predictX = wrapperCF( ua_tr, ua_te, S )
+function [predictX,Su] = wrapperCF( ua_tr, ua_te, S )
 %WRAPPERCF Summary of this function goes here
 %   Detailed explanation goes here
 % ua_tr_tr : Ytrain_train
@@ -6,7 +6,7 @@ function predictX = wrapperCF( ua_tr, ua_te, S )
 % S : Artists_name
 %% top-k count for u users
 k = 5;
-u = [1;109;472;788;982];
+u = [10];
 l = length(u);
 topKL = cell(l+1,k+1);
 for j = 2:k+1
@@ -83,26 +83,27 @@ X = sparse(Su * ua_tr);
 %% Get the score for each 0
 
 umrep = repmat(users_mean,1,size(X,2));
-ummax = repmat(max(ua_tr),1,size(X,2));
-X = X';
+
+
 ua_col= X(:);
 %idx0 = find(ua_col>0);
 %ua_col = ua_col(idx0);
 umrep = umrep(:);
 %umrep = umrep(idx0);
 
-predictX = (ua_col.*ummax).^0.8;
-
+predictX = (ua_col.*umrep).^0.8;
+predictX= reshape(predictX,size(X,1),size(X,2));
+%predictX = (ua_col.*umrep);
 %for i = 1:size(X0,1)
  %   predictX(i,:) = X0(i,:).*users_mean(i);
 %end
 %% compute rmse
-ua_te = ua_te';
+
 ua_tecol = ua_te(:);
 %ua_tecol = ua_tecol(idx0);
 
-rmse = computeCost(ua_tecol,predictX)
-pause;
+rmseTr = rmse(ua_tecol,predictX)
+
 
 %% get the top-k values and their indices
 % Get the top k recommendation for the l users specified at the begining

### 2014-12-14 Project 2/testMyPredSong.m
songPred file

@@ -1,0 +1,51 @@
+ %Store your predictions in a mat file named 'songPred.mat'. 
+% This mat file should contain two matrices 'Ytest_weak_pred'
+% and 'Ytest_strong_pred'. It is clear what names mean.
+% Ytest_weak_pred should be a sparse matrix of size 1774x15082
+% Ytest_stong_pred should be a sparse matrix of size 93x15082
+% One way to create these matrices is shown below:
+%
+%   load songTestPairs;
+%   Ytest_strong_pred = Ytest_strong_pairs;
+%   Ytest_weak_pred = Ytest_weak_pairs;
+%   save('songPred', 'Ytest_strong_pred', 'Ytest_weak_pred');
+%
+% You can then fill in your predictions in the non-zero entries of the two matrices.
+% Once you have the file, running this file will check if the sizes are correct or not.
+
+clear all;
+load songPred; % your file
+load songTestPairs; % test data
+
+% for  Ytest_weak_pred
+% exists?
+if ~exist('Ytest_weak_pred', 'var')
+  error('no Ytest_weak_pred matrix');
+end
+% check size
+[nU,nA] = size(Ytest_weak_pred);
+if nU ~= 1774 & nA ~= 15082
+  error('size of the matrix is not correct');
+end
+% check num of non-zero entries
+if nnz(Ytest_weak_pred) ~= 15924 
+  error('number of non-zero entries is incorrect');
+end
+
+% for Ytest_strong_pred
+% exists?
+if ~exist('Ytest_strong_pred', 'var')
+  error('no Ytest_strong_pred matrix');
+end
+% check size
+[nU,nA] = size(Ytest_strong_pred);
+if nU ~= 93 & nA ~= 15082
+  error('size of the matrix is not correct');
+end
+% check num of non-zero entries
+if nnz(Ytest_strong_pred) ~= 4440
+  error('number of non-zero entries is incorrect');
+end
+
+fprintf('successful, you can submit\n');
+

### 2014-12-14 Project 2/report/paper.tex
report final update

@@
 \subsection{Collaborative-filtering}
-Let $UA$ (users-artists) be equal to $Y_train$.
+Let $UA$ (users-artists) be equal to $Y_{train}$.
 We split the data into a training set $ua_{tr}$ and a test set $ua_{te}$.
@@
 That is because we found the output of NN predictions to be more stable than that of SVM (score between -1 and 1, in contrast to usually off-interval output for SVM).
-The final parameters used are $L_2$ weight = $10^{-6}$, a dropout of 0.43 and dimensions of $[D 250 50 10 2]$, and using sigmood internally, we also increased the number of epochs to 40.
+The final parameters used are $L_2$ weight = $10^{-6}$, a dropout of 0.43 and dimensions of $[D \ 250 \ 50 \ 10 \  2]$, and using sigmood internally, we also increased the number of epochs to 40.\\
+
+For the music recommender we managed to recommend songs successfully to existing users based on their similarity with other users.
+We predicted the counts using different models with sometimes good and very bad results.
+Our best was the model based on the similarity (collaborative-filtering) for the weak prediction.
+Some other model could have been tested if we had more time, for example clustering.
 \subsubsection*{Acknowledgments}

### 2014-12-14 Project 2/report/paper.tex
report final update 2

@@
 \begin{abstract}
-In this project we are given two problems: that of predicting the music listening counts for a given set of users based on their previous listening behaviour and friendship relations; and that of recognizing the presence of people no images.
+In this project we are given two problems: that of predicting the music listening counts for a given set of users based on their previous listening behaviour and friendship relations; and that of recognizing the presence of people in images.
 This report discusses preliminary observations on the given training data, and then results of testing various ML methods for prediction / classification.
@@
 
-\subsection{Penalize Logistic regression (cross-validation for $\lambda$}
+\subsection{Penalize Logistic regression (cross-validation for $\lambda$)}
 Penalized logistic regression was cross-validated (for the lambda parameter) to see which setting gives the best estimated TPR.

### 2014-12-14 Project 2/kevin/mr_mean.m
corrected songPred file and src

@@ -29,7 +29,10 @@ idxw = Ytest_weak_pairs>0;
 idxs = Ytest_strong_pairs>0;
 repm_weak = exp(repm_weak);
 repm_weak(~idxw) = 0;
-
+repm_strong = exp(repm_strong);
+repm_strong(~idxs) = 0;
+Ytest_weak_pred=sparse(repm_weak);
+Ytest_strong_pred = sparse(repm_strong);
 % Ytest_strong_pred(idxs) = repm_strong(idxs);
 
 %%

### 2014-12-14 Project 2/kevin/mr_similarity.m

@@ -56,6 +56,7 @@ Ytest_weak_pred(idxw) = p;
 s = PredStrong(idxs);
 s(s==0)=1;
 Ytest_strong_pred(idxs) = s;
-
+Ytest_weak_pred = exp(Ytest_weak_pred);
+Ytest_strong_pred = exp(Ytest_strong_pred);
 save('songPred', 'Ytest_strong_pred', 'Ytest_weak_pred');
 
